---
title: "pdcl_process_as_1_patient"
output:
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
    toc_depth: 3
    highlight: kate
    fig_width: 10
    fig_heigth: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls() )

# Load required libraries (others are load directly in sourced files)
library(DESeq2)
library(dplyr)
library(purrr)
library(plotly)

# Source files with functions used in next steps
source("R/gprofiler-analysis.R")
source("R/gsea-analysis.R")
source("R/plot-result.R")

# Constant
result_dir <- "~/PhD/PDCL/Result_pdcl_process_as_1_patient/"
```

# Load Data

## Load controls

```{r load-controls}
control_file <- "Data/PDCL/astrocyte_GSE109001_counts_unique_genes_samples_filtered.txt"
control <- read.table(
  control_file, 
  header = T, 
  sep = "\t", 
  as.is = c("id", "symbol")
)
control_names <- c(
  "AF22_NES_Astro_Br1_d29_37_S46",
  "AF22_NES_Astro_Br2_d29_38_S56",
  "AF22_NES_Astro_Br3_d29_39_S66",
  "CCF.STTG1_p24_Br1_S16",
  "CCF.STTG1_p24_Br2_S17",
  "CCF.STTG1_p24_Br3_S18",
  "CDIAstrocytes_p2_Br1_S19",
  "CDIAstrocytes_p2_Br2_S20",
  "CDIAstrocytes_p2_Br3_S21",
  "phaAstrocyte_p2_Br1_S1",
  "phaAstrocyte_p2_Br2_S2",
  "phaAstrocyte_p2_Br3_S3"
  )
```

## Load PDCL DataBase

Load the PDCL count table. Disabled check.names otherwise it will add a X at the beginning of each column name, as column names aren't valid according to R.

```{r load-pdcl}
pdcl_count_table <- read.table(
  "/home/spinicck/PhD/Data/PDCL/lignees_count_genes_PDCL.txt", 
  sep = "\t", 
  header = T, 
  as.is = "gene_id", 
  check.names = F
)

pdcl_names <- c(
  "4339-p21",
  "4371-p37",
  "5706-p14",
  "6190-p43",
  "6240-p12",
  "7015-p17",
  "7060-p18", 
  "7142-p14",
  "N13-1300",
  "N13-1520-p9",
  "N14-0072",
  "N14-0870",
  "N14-1208",
  "N14-1525",
  "N15_0460",
  "N15_0516",
  "N15-0385",
  "N15-0661",
  "N16_0535",
  "N16-0240"
)
```


## Load PENDA results

```{r load-penda}
penda_res_file <- "Data/PDCL/results_combine.csv"
penda_res <- read.table(penda_res_file, sep = ";", header = T, row.names = 1, check.names = F)
penda_pdcl_names <- names(penda_res)
# Remove ".genes.results" string in the column names
penda_pdcl_names <- sub(".genes.results", "", penda_pdcl_names, ignore.case = T)
names(penda_res) <- penda_pdcl_names
```

# Differential Analysis with DESeq2

## Prepare Object needed by DESeq2 for Analysis

Create two data.frame:

-   count_data : contain the raw counts of each genes for both control and PDCL samples
-   count_metadata : metadata for associated with the samples to know which samples belong to the controls and which belong to the PDCL

```{r prepare-count-data}
# Merge control and PDCL counts into one dataframe
count_data <- merge(control, pdcl_count_table, by.x = "symbol", by.y = "gene_id")
# Associate each row with a gene symbol
row.names(count_data) <- count_data$symbol
# Remove symbol and id columns
count_data <- count_data %>% dplyr::select(!c(symbol, id))
condition_vect <- c(
  rep("control", length(control_names)), 
  rep("pdcl", length(pdcl_names))
)
count_metadata <- data.frame(
  sample = c(control_names, pdcl_names), 
  condition=condition_vect
)
count_metadata$condition <- as.factor(count_metadata$condition)
```

## Perform Differential Analysis

Perform a differential analysis for each PDCL in the database against the controls 
and save the result in rds file and a csv file.

```{r run-deseq2, cache=TRUE}
alpha <- 0.05
dds <- DESeqDataSetFromMatrix(
  countData = count_data, 
  colData = count_metadata, 
  design = ~condition
)
dds <- DESeq(dds)
deseq2_res <- results(dds, alpha = alpha, contrast = c("condition", "pdcl", "control"))

saveRDS(deseq2_res, file = paste0(result_dir, "deseq2_pdcl_vs_control.rds"))
write.csv(as.data.frame(deseq2_res), file = paste0(result_dir, "deseq2/deseq2_1_patient.csv"))

deregulated_index <- which(deseq2_res$padj <= alpha)
deregulated_genes <- row.names(deseq2_res)[deregulated_index]

summary(deseq2_res)
```

## VolcanoPlot Differential Analysis Results

```{r volcano-plot-deseq2, echo=FALSE}
ggplot(
  as.data.frame(deseq2_res), 
  aes(x = log2FoldChange, y = -log10(padj), colour = (padj<alpha) )
) + 
  geom_point() +
  scale_color_manual(values = c("TRUE" = "red"), labels = c("Deregulated")) +
  labs(
    title = "Volcano Plot DESeq2 Results",
    colour = NULL
  )
```

# GSEA Pathway Enrichment

## Load Gene Matrix Files (=GMT)

Load GMT files for GSEA pathway enrichment.

```{r create-gsea-gmt-list}
reactome_gmt_file <- gmtPathways(
  "Data/gene-set/jorge-gmt/reactome_gmt_symbol_no_unwanted_categ_no_less_10.gmt"
)
gsea_gmt_list <- list(reactome = reactome_gmt_file)
```

## Run GSEA

Here we run a GSEA pathway enrichment analysis with the log2FoldChange calculated by DESeq2 as the scoring metrics in the ranked list submited to GSEA.

*Message from Jorge :*

We must choose a seed to ensure the analysis is reproducible. This is because GSEA uses a permutation test to span the null distribution of the statistic. Hence, each time we run it, results are expected to change slightly.

```{r run-gsea, cache=TRUE}
set.seed(1)
# Create the ranked list for GSEA
rnk_tmp <- as.data.frame(deseq2_res) %>%
  drop_na() %>%
  filter(padj < alpha) %>%
  mutate(rank = log2FoldChange*-log10(padj)) %>%
  arrange(desc(rank))
rnk <- rnk_tmp %>% pull(rank)
names(rnk) <- row.names(rnk_tmp)
rnk <- rnk[is.finite(rnk)]
rank_list <- list("gbm" = rnk)

# Run GSEA analysis for each ranked list for both Reactom and Bioplanet
gsea_res_list <- lapply(rank_list, run.gsea, gsea_gmt_list)
saveRDS(gsea_res_list, file = paste0(result_dir, "gsea_pathway_enrichment.rds"))

gsea_gem <- lapply(gsea_res_list, convert.gsea.to.gem)
save.gem.list.to.txt(gsea_gem, paste0(result_dir, "gsea/deseq2/"), "deseq2")
```

# GProfiler Pathway Enrichment

Upload GMT File before running G:Profiler Enrichment Analysis if needed.

Otherwise juste define the token of the GMT stored on G:Profiler website.

```{r upload-gmt-gprofiler}
# This portion is commented because the file are already uploaded
# reactome_gmt_token <- upload_GMT_file(gmtfile = 
#  "Data/gene-set/jorge-gmt/reactome_gmt_symbol_no_unwanted_categ_no_less_10.gmt"
# )
# bioplanet_gmt_token <- upload_GMT_file(gmtfile = 
#  "Data/gene-set/jorge-gmt/bioplanet_gmt_symbol_no_unwanted_categ_no_less_10.gmt"
# )

reactome_gmt_token <- "gp__Nqmx_yxm7_37U"
bioplanet_gmt_token <- "gp__vB7V_DT57_yNY"
```

Create a list of GMT files.

```{r create-gprofiler-gmt-list}
gprofiler_gmt_list <- list(reactome = reactome_gmt_token)
```

Run G:Profiler for PENDA result

```{r run-gprofiler, cache=TRUE}
all_pdcl_genes <- row.names(deseq2_res)
all_gene_list <- list("gbm" = deregulated_genes)
gost_res_list <- lapply(all_gene_list, function(gene_list){
  res <- run.gost(
    gene_list,
    all_pdcl_genes,
    gprofiler_gmt_list
  )
  return(res)
})
```

Save Result in RDS and in Generic Enrichment Map (GEM) format

```{r save-gprofiler}
saveRDS(gost_res_list, file = paste0(result_dir, "gprofiler_pathway_enrichment.rds"))
gost_gem <- lapply(gost_res_list, convert.gost.to.gem)
save.gem.list.to.txt(gost_gem, paste0(result_dir, "gprofiler/deseq2/"), "deseq2")
```

# Process Result


## Define GEM list

```{r define-gem-list}
gem_list <- list("G:Profiler" = gost_gem, "GSEA" = gsea_gem)
```

## Collapse by method

```{r collapse-by-method}
# Collapse the results by method.
# One line contain the result of one enrichment method for one pathway of one PDCL
gem_collapsed <- collapse.gem.list(gem_list)
```

## Collapse by pathways

```{r collapse-by-pathways}
gem_collapsed_method_vs_method <- collapse.method.vs.method(gem_collapsed)
```

## Find common pathways

```{r find-common-pathways}
all_common_pathways <- find.method.common.pathways(gem_collapsed_method_vs_method, c(0.05, 0.1))
gem_collapsed_method_vs_method <- gem_collapsed_method_vs_method %>%
  inner_join(all_common_pathways, by=c("pathway_id", "pdcl", "method_1", "method_2"))
```

# Plot 

```{r set-alpha, include=FALSE}
threshold <- 0.1
plot_comp_method_dir <- paste0(result_dir, "plot/alpha_0.05/comparison_method/")
count_dir <- paste0(result_dir, "plot/alpha_0.05/count/")
heatmap_dir <- paste0(result_dir, "plot/alpha_0.05/heatmap/")
```

## Enriched Pathways

```{r enriched-pathways}
gem_collapsed %>%
  filter(fdr<threshold) %>%
  count(method)
```

## Plot count categories 

```{r count-categories-10-percent}
count_categories_0.1 <- plot.count.categories(gem_collapsed, threshold)
tmp_filename <- paste0(count_dir, "plot_count_categories_0.1.png")
ggsave(tmp_filename, plot = count_categories_0.1, width = 10, height = 7)
count_categories_0.1
```

```{r save-session, include=FALSE}
save.image("~/PhD/PDCL/pdcl_process_as_1_patient.RData")
```

