---
title: "PDCL Process"
date: "`r Sys.Date()`"
author: "Kévin SPINICCI"
output:
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
    toc_depth: 3
    highlight: kate
    fig_width: 10
    fig_heigth: 7
---


```{r setup-library, include=FALSE}
# clean workspace
rm(list=ls()) 

# Load required libraries (others are load directly in sourced files)
library(DESeq2)
library(dplyr)
library(purrr)

# Source files with functions used in next steps
source("R/pdcl_process_personalized_for_each_patient/gprofiler-analysis.R")
source("R/pdcl_process_personalized_for_each_patient/gsea-analysis.R")
source("R/pdcl_process_personalized_for_each_patient/plot-result.R")
```

# Load Data

## Load controls

```{r load-controls}
control_file <- "Data/PDCL/astrocyte_GSE109001_counts_unique_genes_samples_filtered.txt"
control <- read.table(
  control_file, 
  header = T, 
  sep = "\t", 
  as.is = c("id", "symbol")
)
control_names <- c(
  "AF22_NES_Astro_Br1_d29_37_S46",
  "AF22_NES_Astro_Br2_d29_38_S56",
  "AF22_NES_Astro_Br3_d29_39_S66",
  "CCF.STTG1_p24_Br1_S16",
  "CCF.STTG1_p24_Br2_S17",
  "CCF.STTG1_p24_Br3_S18",
  "CDIAstrocytes_p2_Br1_S19",
  "CDIAstrocytes_p2_Br2_S20",
  "CDIAstrocytes_p2_Br3_S21",
  "phaAstrocyte_p2_Br1_S1",
  "phaAstrocyte_p2_Br2_S2",
  "phaAstrocyte_p2_Br3_S3"
  )
```

## Load PDCL DataBase

Load the PDCL count table. Disabled check.names otherwise it will add a X at the beginning of each column name, as column names aren't valid according to R.

```{r load-pdcl}
pdcl_count_table <- read.table(
  "/home/spinicck/PhD/Data/PDCL/lignees_count_genes_PDCL.txt", 
  sep = "\t", 
  header = T, 
  as.is = "gene_id", 
  check.names = F
)

pdcl_names <- c(
  "4339-p21",
  "4371-p37",
  "5706-p14",
  "6190-p43",
  "6240-p12",
  "7015-p17",
  "7060-p18", 
  "7142-p14",
  "N13-1300",
  "N13-1520-p9",
  "N14-0072",
  "N14-0870",
  "N14-1208",
  "N14-1525",
  "N15_0460",
  "N15_0516",
  "N15-0385",
  "N15-0661",
  "N16_0535",
  "N16-0240"
)
```


## Load PENDA results

```{r load-penda}
penda_res_file <- "Data/PDCL/results_combine.csv"
penda_res <- read.table(penda_res_file, sep = ";", header = T, row.names = 1, check.names = F)
penda_pdcl_names <- names(penda_res)
# Remove ".genes.results" string in the column names
penda_pdcl_names <- sub(".genes.results", "", penda_pdcl_names, ignore.case = T)
names(penda_res) <- penda_pdcl_names
```

# Differential Analysis with DESeq2

## Prepare Object needed by DESeq2 for Analysis

Create two data.frame:

-   count_data : contain the raw counts of each genes for both control and PDCL samples
-   count_metadata : metadata for associated with the samples to know which samples belong to the controls and which belong to the PDCL

```{r prepare-count-data}
# Merge control and PDCL counts into one dataframe
count_data <- merge(control, pdcl_count_table, by.x = "symbol", by.y = "gene_id")
# Associate each row with a gene symbol
row.names(count_data) <- count_data$symbol
# Remove symbol and id columns
count_data <- count_data %>% dplyr::select(!c(symbol, id))
condition_vect <- c(
  rep("control", length(control_names)), 
  rep("pdcl", length(pdcl_names))
)
count_metadata <- data.frame(
  sample = c(control_names, pdcl_names), 
  condition=condition_vect
)
count_metadata$condition <- as.factor(count_metadata$condition)
```

## Perform Differential Analysis

Perform a differential analysis for each PDCL in the database against the controls.

```{r run-deseq2, cache=TRUE}
deseq2_res_list <- lapply(pdcl_names, function(pdcl){
  dataset_pdcl <- c(control_names, pdcl)
  # Subract a dataset which contains only the counts for the
  # current PDCL
  dataset <- count_data[,dataset_pdcl]
  # Same with metadata
  dataset_metadata <- count_metadata[ count_metadata$sample %in% dataset_pdcl, ]
  # Perform DESeq2 Analysis
  dds <- DESeqDataSetFromMatrix(
    countData = dataset, 
    colData = dataset_metadata, 
    design = ~condition
  )
  dds <- DESeq(dds)
  res <- results(dds, alpha = 0.05)
})
names(deseq2_res_list) <- pdcl_names
```

Save Result in rds file and write a csv file for each PDCL.

```{r save-deseq2}
saveRDS(deseq2_res_list, file = "Result/deseq2_pdcl_vs_control.rds")
for (pdcl in pdcl_names){
  res <- deseq2_res_list[[pdcl]]
  res.file <- paste0("Result/deseq2/deseq2_", pdcl, ".csv")
  write.csv(as.data.frame(res), file = res.file)
}
```

## DESeq2 comparison with PENDA

Here we compare the sign of the log2FoldChange given by DESeq2 with the deregulation
predicted by PENDA.

```{r comp-deseq2-and-penda, cache=TRUE, fig.width=16, fig.height=10}
comp_deseq2_penda <- lapply(pdcl_names, function(pdcl){
  deseq2 <- as.data.frame(deseq2_res_list[[pdcl]])[,c("log2FoldChange", "padj")]
  x <- merge(deseq2, penda_res, by = "row.names")
  x <- x[,c("Row.names", "log2FoldChange", "padj", pdcl)]
  names(x) <- c("gene", "log2FoldChange", "padj", "penda")
  x
})
names(comp_deseq2_penda) <- pdcl_names

lapply(comp_deseq2_penda, function(x){
  x$deseq2 <- sign(x$log2FoldChange)
  addmargins(table(x[,c("deseq2", "penda")]))
})

volcano_comp_deseq2_penda <- volcano.deseq2.vs.penda.all(deseq2_res_list, penda_res, pdcl_names)
ggsave(
  "Result/plot/volcano_comp_deseq2_penda.png",
  plot = volcano_comp_deseq2_penda,
  width = 16,
  height = 10
)
volcano_comp_deseq2_penda
```

# GSEA Pathway Enrichment

## Load Gene Matrix Files (=GMT)

Load GMT files for GSEA pathway enrichment.

```{r create-gsea-gmt-list}
reactome_gmt_file <- gmtPathways(
  "Data/gene-set/jorge-gmt/reactome_gmt_symbol_no_unwanted_categ_no_less_10.gmt"
)
gsea_gmt_list <- list(reactome = reactome_gmt_file)
```

## Run GSEA

Here we run a GSEA pathway enrichment analysis with the log2FoldChange calculated by DESeq2 as the scoring metrics in the ranked list submited to GSEA.

*Message from Jorge :*

We must choose a seed to ensure the analysis is reproducible. This is because GSEA uses a permutation test to span the null distribution of the statistic. Hence, each time we run it, results are expected to change slightly.

```{r run-gsea, cache=TRUE}
set.seed(1)
# Create the ranked list for GSEA
rank_list <- lapply(deseq2_res_list, function(x){
  rnk <- x$log2FoldChange
  names(rnk) <- row.names(x)
  rnk <- sort(rnk, decreasing = T)
  rnk
})

# Run GSEA analysis for each ranked list for both Reactom and Bioplanet
gsea_res_list <- lapply(rank_list, run.gsea, gsea_gmt_list)
saveRDS(gsea_res_list, file = "Result/gsea_pathway_enrichment.rds")

gsea_gem <- lapply(gsea_res_list, convert.gsea.to.gem)
save.gem.list.to.txt(gsea_gem, "Result/gsea/deseq2/", "deseq2")
```

## Run GSEA with processing using PENDA

Here we run a GSEA pathway enrichment analysis with the log2FoldChange calculated by DESeq2 as the scoring metrics in the ranked list submited to GSEA.
We use the results from the PENDA differential expression analysis to set to 0 the rank for genes that are not found differentially expressed by PENDA. 
We cannot trust the DESeq2 p-value as the condition for the test were not satisfied (multiple replicates for each PDCL).

*Message from Jorge :*

We must choose a seed to ensure the analysis is reproducible. This is because GSEA uses a permutation test to span the null distribution of the statistic. Hence, each time we run it, results are expected to change slightly.

```{r run-gsea-processed, cache=TRUE}
set.seed(1)

penda_genes <- row.names(penda_res)
rank_list_procesed <- lapply(comp_deseq2_penda, function(x){
  x <- x[!is.na(x$padj),]
  keep <- which( sign(x$log2FoldChange) == x$penda)
  rnk <- x$log2FoldChange[keep]
  names(rnk) <- x$gene[keep]
  rnk <- sort(rnk, decreasing = T)
  rnk
})

gsea_processed_res_list <- lapply(rank_list_procesed, run.gsea, gsea_gmt_list)
saveRDS(gsea_processed_res_list, file = "Result/gsea_processed_pathway_enrichment.rds")

gsea_processed_gem <- lapply(gsea_processed_res_list, convert.gsea.to.gem)
save.gem.list.to.txt(gsea_processed_gem, "Result/gsea_processed/deseq2/", "deseq2")
```

# GProfiler Analysis for PENDA resuls

Upload GMT File before running G:Profiler Enrichment Analysis if needed.

Otherwise juste define the token of the GMT stored on G:Profiler website.

```{r upload-gmt-gprofiler}
# This portion is commented because the file are already uploaded
# reactome_gmt_token <- upload_GMT_file(gmtfile = 
#  "Data/gene-set/jorge-gmt/reactome_gmt_symbol_no_unwanted_categ_no_less_10.gmt"
# )
# bioplanet_gmt_token <- upload_GMT_file(gmtfile = 
#  "Data/gene-set/jorge-gmt/bioplanet_gmt_symbol_no_unwanted_categ_no_less_10.gmt"
# )

reactome_gmt_token <- "gp__Nqmx_yxm7_37U"
bioplanet_gmt_token <- "gp__vB7V_DT57_yNY"
```

Create a list of GMT files.

```{r create-gprofiler-gmt-list}
gprofiler_gmt_list <- list(reactome = reactome_gmt_token)
```

Run G:Profiler for PENDA result

```{r run-gprofiler, cache=TRUE}
all_pdcl_genes <- row.names(penda_res)
gost_res_list <- lapply(penda_res, function(pdcl){
  gene_list <- all_pdcl_genes[pdcl != 0]
  res <- run.gost(
    gene_list,
    all_pdcl_genes,
    gprofiler_gmt_list
  )
  return(res)
})
```

Save Result in RDS and in Generic Enrichment Map (GEM) format

```{r save-gprofiler}
saveRDS(gost_res_list, file = "Result/gprofiler_pathway_enrichment.rds")
gost_gem <- lapply(gost_res_list, convert.gost.to.gem)
save.gem.list.to.txt(gost_gem, "Result/gprofiler/penda/", "penda")
```

# Process Result

## Define GEM list

```{r define-gem-list}
gem_list <- list("G:Profiler" = gost_gem, "GSEA" = gsea_gem, "GSEA Processed" = gsea_processed_gem)
```

## Collapse by method

```{r collapse-by-method}
# Collapse the results by method.
# One line contain the result of one enrichment method for one pathway of one PDCL
gem_collapsed <- collapse.gem.list(gem_list)
```

## Collapse by pathways

```{r collapse-by-pathways}
gem_collapsed_method_vs_method <- collapse.method.vs.method(gem_collapsed)
```

## Find common pathways

```{r find-common-pathways}
all_common_pathways <- find.method.common.pathways(gem_collapsed_method_vs_method, c(0.05, 0.1))
gem_collapsed_method_vs_method <- gem_collapsed_method_vs_method %>%
  inner_join(all_common_pathways, by=c("pathway_id", "pdcl", "method_1", "method_2"))
```

## Collapse rank list

```{r collapse-rank-list}
rnk_collapsed <- collapse.rnk(rank_list)
rnk_processed_collapsed <- collapse.rnk(rank_list_procesed)
```

# Plot leading edge for GSEA

```{r plot-gsea-leading-edge-10-percent, fig.width=24, fig.height=8}
gsea_leading_edge_0.1 <- lapply(pdcl_names, function(pdcl){
  plot_tmp <- plot.leading.edge(rnk_collapsed, gem_collapsed, pdcl, 0.1)
  filepath <- paste0("Result/plot/alpha_0.1/leadingEdge/gsea/", pdcl, "_gsea_leadingedge.png")
  ggsave(filepath, plot = plot_tmp, width = 24, height = 8)
  plot_tmp
})
names(gsea_leading_edge_0.1) <- pdcl_names
gsea_leading_edge_0.1
```

# Plot leading edge for GSEA Processed

```{r plot-gsea-processed-leading-edge-10-percent, fig.width=24, fig.height=8}
gsea_processed_leading_edge_0.1 <- lapply(pdcl_names, function(pdcl){
  plot_tmp <- plot.leading.edge(rnk_processed_collapsed, gem_collapsed, pdcl, 0.1, "GSEA Processed")
  filepath <- paste0("Result/plot/alpha_0.1/leadingEdge/gsea_processed/", pdcl, "_gsea_processed_leadingedge.png")
  ggsave(filepath, plot = plot_tmp, width = 24, height = 8)
  plot_tmp
})
names(gsea_processed_leading_edge_0.1) <- pdcl_names
gsea_processed_leading_edge_0.1
```

# Plot Results for alpha = 0.05

```{r set-alpha-5-percent}
threshold <- 0.05
plot_comp_method_dir <- "Result/plot/alpha_0.05/comparison_method/"
count_dir <- "Result/plot/alpha_0.05/count/"
heatmap_dir <- "Result/plot/alpha_0.05/heatmap/"
```

### Plot Method vs Method 

#### G:Profiler vs GSEA
```{r gprofiler-vs-gsea-5-percent}
gprofiler_vs_gsea <- plot_method_vs_method(gem_collapsed_method_vs_method, "G:Profiler", "GSEA", threshold)
tmp_filename <- paste0(plot_comp_method_dir, "plot_gprofiler_vs_gsea.png")
ggsave(tmp_filename, plot = gprofiler_vs_gsea, width = 10, height = 7)
gprofiler_vs_gsea
```


#### G:Profiler vs GSEA Processed
```{r gprofiler-vs-gsea-processed-5-percent}
gprofiler_vs_gsea_processed <- plot_method_vs_method(gem_collapsed_method_vs_method, "G:Profiler", "GSEA Processed", threshold)
tmp_filename <- paste0(plot_comp_method_dir, "plot_gprofiler_vs_gsea_processed.png")
ggsave(tmp_filename, plot = gprofiler_vs_gsea_processed, width = 10, height = 7)
gprofiler_vs_gsea_processed
```

#### GSEA vs GSEA Processed
```{r gsea-vs-gsea-processed-5-percent}
gsea_vs_gsea_processed <- plot_method_vs_method(gem_collapsed_method_vs_method, "GSEA", "GSEA Processed", threshold)
tmp_filename <- paste0(plot_comp_method_dir, "plot_gsea_vs_gsea_processed.png")
ggsave(tmp_filename, plot = gsea_vs_gsea_processed, width = 10, height = 7)
gsea_vs_gsea_processed
```

### Plot count categories 


#### All method
```{r count-categories-5-percent}
count_categories_0.05 <- plot.count.categories(gem_collapsed, threshold)
tmp_filename <- paste0(count_dir, "plot_count_categories_0.05.png")
ggsave(tmp_filename, plot = count_categories_0.05, width = 10, height = 7)
count_categories_0.05
```

#### G:Profiler vs GSEA
```{r count-categories-gprofiler-vs-gsea-5-percent}

count_categories_gprofiler_vs_gsea_0.05 <- plot.count.categories.m.vs.m(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA")
tmp_filename <- paste0(count_dir, "plot_count_categories_gprofiler_vs_gsea_0.05.png")
ggsave(tmp_filename, plot = count_categories_gprofiler_vs_gsea_0.05, width = 10, height = 7)
count_categories_gprofiler_vs_gsea_0.05
```

#### G:Profiler vs GSEA Processed
```{r count-categories-gprofiler-vs-gsea_processed-5-percent}

count_categories_gprofiler_vs_gsea_processed_0.05 <- plot.count.categories.m.vs.m(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA Processed")
tmp_filename <- paste0(count_dir, "plot_count_categories_gprofiler_vs_gsea_processed_0.05.png")
ggsave(tmp_filename, plot = count_categories_gprofiler_vs_gsea_processed_0.05, width = 10, height = 7)
count_categories_gprofiler_vs_gsea_processed_0.05
```

#### GSEA vs GSEA Processed
```{r count-categories-gsea-vs-gsea_processed-5-percent}

count_categories_gsea_vs_gsea_processed_0.05 <- plot.count.categories.m.vs.m(gem_collapsed, all_common_pathways, threshold, "GSEA", "GSEA Processed")
tmp_filename <- paste0(count_dir, "plot_count_categories_gsea_vs_gsea_processed_0.05.png")
ggsave(tmp_filename, plot = count_categories_gsea_vs_gsea_processed_0.05, width = 10, height = 7)
count_categories_gsea_vs_gsea_processed_0.05
```

### Plot count pathways 

#### All Method

```{r count-pathways-5-percent}
count_pathways_0.05 <- plot.count.pathways(gem_collapsed, threshold)
tmp_filename <- paste0(count_dir, "plot_count_pathways.png")
ggsave(tmp_filename, plot = count_pathways_0.05, width = 10, height = 7)
count_pathways_0.05 
```

#### G:Profiler vs GSEA

```{r count-pathways-gprofiler-vs-gsea-5-percent}
count_pathways_gprofiler_vs_gsea_0.05 <- plot.count.pathways.m.vs.m(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA")
tmp_filename <- paste0(count_dir, "plot_count_pathways_gprofiler_vs_gsea.png")
ggsave(tmp_filename, plot = count_pathways_gprofiler_vs_gsea_0.05, width = 10, height = 7)
count_pathways_gprofiler_vs_gsea_0.05 
```

#### G:Profiler vs GSEA Processed

```{r count-pathways-gprofiler-vs-gsea_processed-5-percent}
count_pathways_gprofiler_vs_gsea_processed_0.05 <- plot.count.pathways.m.vs.m(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA Processed")
tmp_filename <- paste0(count_dir, "plot_count_pathways_gprofiler_vs_gsea_processed.png")
ggsave(tmp_filename, plot = count_pathways_gprofiler_vs_gsea_processed_0.05, width = 10, height = 7)
count_pathways_gprofiler_vs_gsea_processed_0.05 
```

#### GSEA vs GSEA Processed

```{r count-pathways-gsea-vs-gsea_processed-5-percent}
count_pathways_gsea_vs_gsea_processed_0.05 <- plot.count.pathways.m.vs.m(gem_collapsed, all_common_pathways, threshold, "GSEA", "GSEA Processed")
tmp_filename <- paste0(count_dir, "plot_count_pathways_gsea_vs_gsea_processed.png")
ggsave(tmp_filename, plot = count_pathways_gsea_vs_gsea_processed_0.05, width = 10, height = 7)
count_pathways_gsea_vs_gsea_processed_0.05 
```

### Plot Heatmap FDR 

Here we plot a heatmap of the fdr values of a pathways for each PDCL for all methods.
Black rectangles indicates pathways that are enriched in two methods selected for comparison.

#### G:Profiler vs GSEA

```{r heatmap-pathways-gprofiler-vs-gsea-5-percent, fig.width=16, fig.height=10}
heatmap_pathways_gprofiler_vs_gsea_0.05 <- heatmap.pathways(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA")
tmp_filename <- paste0(heatmap_dir, "heatmap_pathways_gprofiler_vs_gsea.png")
ggsave(tmp_filename, plot = heatmap_pathways_gprofiler_vs_gsea_0.05, width = 16, height = 10)
heatmap_pathways_gprofiler_vs_gsea_0.05
```

#### G:Profiler vs GSEA Processed

```{r heatmap-pathways-gprofiler-vs-gsea_processed-5-percent, fig.width=16, fig.height=10}
heatmap_pathways_gprofiler_vs_gsea_processed_0.05 <- heatmap.pathways(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA Processed")
tmp_filename <- paste0(heatmap_dir, "heatmap_pathways_gprofiler_vs_gsea_processed.png")
ggsave(tmp_filename, plot = heatmap_pathways_gprofiler_vs_gsea_processed_0.05, width = 16, height = 10)
heatmap_pathways_gprofiler_vs_gsea_processed_0.05
```

#### GSEA vs GSEA Processed

```{r heatmap-pathways-gsea-vs-gsea_processed-5-percent,  fig.width=16, fig.height=10}
heatmap_pathways_gsea_vs_gsea_processed_0.05 <- heatmap.pathways(gem_collapsed, all_common_pathways, threshold, "GSEA", "GSEA Processed")
tmp_filename <- paste0(heatmap_dir, "heatmap_pathways_gsea_vs_gsea_processed.png")
ggsave(tmp_filename, plot = heatmap_pathways_gsea_vs_gsea_processed_0.05, width = 16, height = 10)
heatmap_pathways_gsea_vs_gsea_processed_0.05
```

### Plot heatmap Categories

```{r heatmap-categories-5-percent, fig.width=16, fig.height=10}
heatmap_categories_0.05 <- heatmap.categories(gem_collapsed, threshold)
tmp_filename <- paste0(heatmap_dir, "heatmap_categories.png")
ggsave(tmp_filename, plot = heatmap_categories_0.05, width = 16, height = 10)
heatmap_categories_0.05
```

# Plot Results for alpha = 0.1

```{r set-alpha-10-percent}
threshold <- 0.1
plot_comp_method_dir <- "Result/plot/alpha_0.1/comparison_method/"
count_dir <- "Result/plot/alpha_0.1/count/"
heatmap_dir <- "Result/plot/alpha_0.1/heatmap/"
```

### Plot Method vs Method 

#### G:Profiler vs GSEA
```{r gprofiler-vs-gsea-10-percent}
gprofiler_vs_gsea <- plot_method_vs_method(gem_collapsed_method_vs_method, "G:Profiler", "GSEA", threshold)
tmp_filename <- paste0(plot_comp_method_dir, "plot_gprofiler_vs_gsea.png")
ggsave(tmp_filename, plot = gprofiler_vs_gsea, width = 10, height = 7)
gprofiler_vs_gsea
```


#### G:Profiler vs GSEA Processed
```{r gprofiler-vs-gsea-processed-10-percent}
gprofiler_vs_gsea_processed <- plot_method_vs_method(gem_collapsed_method_vs_method, "G:Profiler", "GSEA Processed", threshold)
tmp_filename <- paste0(plot_comp_method_dir, "plot_gprofiler_vs_gsea_processed.png")
ggsave(tmp_filename, plot = gprofiler_vs_gsea_processed, width = 10, height = 7)
gprofiler_vs_gsea_processed
```

#### GSEA vs GSEA Processed
```{r gsea-vs-gsea-processed-10-percent}
gsea_vs_gsea_processed <- plot_method_vs_method(gem_collapsed_method_vs_method, "GSEA", "GSEA Processed", threshold)
tmp_filename <- paste0(plot_comp_method_dir, "plot_gsea_vs_gsea_processed.png")
ggsave(tmp_filename, plot = gsea_vs_gsea_processed, width = 10, height = 7)
gsea_vs_gsea_processed
```

### Plot count categories 


#### All method
```{r count-categories-10-percent}
count_categories_0.1 <- plot.count.categories(gem_collapsed, threshold)
tmp_filename <- paste0(count_dir, "plot_count_categories_0.1.png")
ggsave(tmp_filename, plot = count_categories_0.1, width = 10, height = 7)
count_categories_0.1
```

#### G:Profiler vs GSEA
```{r count-categories-gprofiler-vs-gsea-10-percent}

count_categories_gprofiler_vs_gsea_0.1 <- plot.count.categories.m.vs.m(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA")
tmp_filename <- paste0(count_dir, "plot_count_categories_gprofiler_vs_gsea_0.1.png")
ggsave(tmp_filename, plot = count_categories_gprofiler_vs_gsea_0.1, width = 10, height = 7)
count_categories_gprofiler_vs_gsea_0.1
```

#### G:Profiler vs GSEA Processed
```{r count-categories-gprofiler-vs-gsea_processed-10-percent}

count_categories_gprofiler_vs_gsea_processed_0.1 <- plot.count.categories.m.vs.m(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA Processed")
tmp_filename <- paste0(count_dir, "plot_count_categories_gprofiler_vs_gsea_processed_0.1.png")
ggsave(tmp_filename, plot = count_categories_gprofiler_vs_gsea_processed_0.1, width = 10, height = 7)
count_categories_gprofiler_vs_gsea_processed_0.1
```

#### GSEA vs GSEA Processed
```{r count-categories-gsea-vs-gsea_processed-10-percent}

count_categories_gsea_vs_gsea_processed_0.1 <- plot.count.categories.m.vs.m(gem_collapsed, all_common_pathways, threshold, "GSEA", "GSEA Processed")
tmp_filename <- paste0(count_dir, "plot_count_categories_gsea_vs_gsea_processed_0.1.png")
ggsave(tmp_filename, plot = count_categories_gsea_vs_gsea_processed_0.1, width = 10, height = 7)
count_categories_gsea_vs_gsea_processed_0.1
```

### Plot count pathways 

#### All Method

```{r count-pathways-10-percent}
count_pathways_0.1 <- plot.count.pathways(gem_collapsed, threshold)
tmp_filename <- paste0(count_dir, "plot_count_pathways.png")
ggsave(tmp_filename, plot = count_pathways_0.1, width = 10, height = 7)
count_pathways_0.1 
```

#### G:Profiler vs GSEA

```{r count-pathways-gprofiler-vs-gsea-10-percent}
count_pathways_gprofiler_vs_gsea_0.1 <- plot.count.pathways.m.vs.m(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA")
tmp_filename <- paste0(count_dir, "plot_count_pathways_gprofiler_vs_gsea.png")
ggsave(tmp_filename, plot = count_pathways_gprofiler_vs_gsea_0.1, width = 10, height = 7)
count_pathways_gprofiler_vs_gsea_0.1 
```

#### G:Profiler vs GSEA Processed

```{r count-pathways-gprofiler-vs-gsea_processed-10-percent}
count_pathways_gprofiler_vs_gsea_processed_0.1 <- plot.count.pathways.m.vs.m(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA Processed")
tmp_filename <- paste0(count_dir, "plot_count_pathways_gprofiler_vs_gsea_processed.png")
ggsave(tmp_filename, plot = count_pathways_gprofiler_vs_gsea_processed_0.1, width = 10, height = 7)
count_pathways_gprofiler_vs_gsea_processed_0.1 
```

#### GSEA vs GSEA Processed

```{r count-pathways-gsea-vs-gsea_processed-10-percent}
count_pathways_gsea_vs_gsea_processed_0.1 <- plot.count.pathways.m.vs.m(gem_collapsed, all_common_pathways, threshold, "GSEA", "GSEA Processed")
tmp_filename <- paste0(count_dir, "plot_count_pathways_gsea_vs_gsea_processed.png")
ggsave(tmp_filename, plot = count_pathways_gsea_vs_gsea_processed_0.1, width = 10, height = 7)
count_pathways_gsea_vs_gsea_processed_0.1 
```

### Plot Heatmap FDR 

Here we plot a heatmap of the fdr values of a pathways for each PDCL for all methods.
Black rectangles indicates pathways that are enriched in two methods selected for comparison.

#### G:Profiler vs GSEA

```{r heatmap-pathways-gprofiler-vs-gsea-10-percent, fig.width=16, fig.height=10}
heatmap_pathways_gprofiler_vs_gsea_0.1 <- heatmap.pathways(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA")
tmp_filename <- paste0(heatmap_dir, "heatmap_pathways_gprofiler_vs_gsea.png")
ggsave(tmp_filename, plot = heatmap_pathways_gprofiler_vs_gsea_0.1, width = 16, height = 10)
heatmap_pathways_gprofiler_vs_gsea_0.1
```

#### G:Profiler vs GSEA Processed

```{r heatmap-pathways-gprofiler-vs-gsea_processed-10-percent, fig.width=16, fig.height=10}
heatmap_pathways_gprofiler_vs_gsea_processed_0.1 <- heatmap.pathways(gem_collapsed, all_common_pathways, threshold, "G:Profiler", "GSEA Processed")
tmp_filename <- paste0(heatmap_dir, "heatmap_pathways_gprofiler_vs_gsea_processed.png")
ggsave(tmp_filename, plot = heatmap_pathways_gprofiler_vs_gsea_processed_0.1, width = 16, height = 10)
heatmap_pathways_gprofiler_vs_gsea_processed_0.1
```

#### GSEA vs GSEA Processed

```{r heatmap-pathways-gsea-vs-gsea_processed-10-percent,  fig.width=16, fig.height=10}
heatmap_pathways_gsea_vs_gsea_processed_0.1 <- heatmap.pathways(gem_collapsed, all_common_pathways, threshold, "GSEA", "GSEA Processed")
tmp_filename <- paste0(heatmap_dir, "heatmap_pathways_gsea_vs_gsea_processed.png")
ggsave(tmp_filename, plot = heatmap_pathways_gsea_vs_gsea_processed_0.1, width = 16, height = 10)
heatmap_pathways_gsea_vs_gsea_processed_0.1
```

### Plot heatmap Categories

```{r heatmap-categories-10-percent, fig.width=16, fig.height=10}
heatmap_categories_0.1 <- heatmap.categories(gem_collapsed, threshold)
tmp_filename <- paste0(heatmap_dir, "heatmap_categories.png")
ggsave(tmp_filename, plot = heatmap_categories_0.1, width = 16, height = 10)
heatmap_categories_0.1
```

```{r save-workspace, include=FALSE}
save.image("~/PhD/PDCL/pdcl.RData")
```

