---
title: "pdcl_process_as_1_patient"
output:
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
    toc_depth: 3
    highlight: kate
    fig_width: 10
    fig_heigth: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls() )

# Load required libraries (others are load directly in sourced files)
library(DESeq2)
library(dplyr)
library(purrr)
library(plotly)
library(DT)

# Source files with functions used in next steps
source("R/pdcl_process_as_1_patient/gprofiler-analysis.R")
source("R/pdcl_process_as_1_patient/gsea-analysis.R")
source("R/pdcl_process_as_1_patient/process-result.R")
source("R/pdcl_process_as_1_patient/guickgo.R")

# Constant
result_dir <- "~/PhD/PDCL/Result_pdcl_process_as_1_patient/"
```

# Load Data

## Load controls

```{r load-controls}
control_file <- "Data/PDCL/astrocyte_GSE109001_counts_unique_genes_samples_filtered.txt"
control <- read.table(
  control_file, 
  header = T, 
  sep = "\t", 
  as.is = c("id", "symbol")
)
control_names <- c(
  "AF22_NES_Astro_Br1_d29_37_S46",
  "AF22_NES_Astro_Br2_d29_38_S56",
  "AF22_NES_Astro_Br3_d29_39_S66",
  "CCF.STTG1_p24_Br1_S16",
  "CCF.STTG1_p24_Br2_S17",
  "CCF.STTG1_p24_Br3_S18",
  "CDIAstrocytes_p2_Br1_S19",
  "CDIAstrocytes_p2_Br2_S20",
  "CDIAstrocytes_p2_Br3_S21",
  "phaAstrocyte_p2_Br1_S1",
  "phaAstrocyte_p2_Br2_S2",
  "phaAstrocyte_p2_Br3_S3"
  )
```

## Load PDCL DataBase

Load the PDCL count table. Disabled check.names otherwise it will add a X at the beginning of each column name, as column names aren't valid according to R.

```{r load-pdcl}
pdcl_count_table <- read.table(
  "/home/spinicck/PhD/Data/PDCL/lignees_count_genes_PDCL.txt", 
  sep = "\t", 
  header = T, 
  as.is = "gene_id", 
  check.names = F
)

pdcl_names <- c(
  "4339-p21",
  "4371-p37",
  "5706-p14",
  "6190-p43",
  "6240-p12",
  "7015-p17",
  "7060-p18", 
  "7142-p14",
  "N13-1300",
  "N13-1520-p9",
  "N14-0072",
  "N14-0870",
  "N14-1208",
  "N14-1525",
  "N15_0460",
  "N15_0516",
  "N15-0385",
  "N15-0661",
  "N16_0535",
  "N16-0240"
)
```


## Load PENDA results

```{r load-penda}
penda_res <- readRDS("Data/PDCL/results_penda_astrocytes_2021.rds")
penda_res <- as.data.frame(penda_res)
penda_pdcl_names <- names(penda_res)
# Remove ".genes.results" string in the column names
penda_pdcl_names <- sub(".genes.results", "", penda_pdcl_names, ignore.case = T)
names(penda_res) <- penda_pdcl_names
```

## Define GMT path
```{r define-gmt-path}
gmt_path <- list(
  reactome = "Data/gene-set/reactome_no_less_15_no_more_1000.gmt",
  kegg = "Data/gene-set/kegg_no_less_15_no_more_1000.gmt"
)
```

## Load GSEA GMTs

Load GMT files for GSEA pathway enrichment.

```{r create-gsea-gmt-list}
gsea_gmt_list <- lapply(gmt_path, function(x){ gmtPathways(x)})
```

## Load G:Profiler GMTs

Upload GMT File before running G:Profiler Enrichment Analysis if needed.
Otherwise juste define the token of the GMT stored on G:Profiler website.

```{r create-gprofiler-gmt-list}
# This portion is commented because the file are already uploaded
gprofiler_gmt_list <- lapply(gmt_path, function(x){ upload_GMT_file(x) })
# bioplanet_gmt_token <- "gp__vB7V_DT57_yNY"
# gprofiler_gmt_list <- list(reactome = "gp__Nqmx_yxm7_37U", kegg = "gp__5f4Q_8t3p_PVw")
# gprofiler_gmt_list <- list(reactome = "gp__VxvL_Ua0o_Irk", kegg = "gp__5f4Q_8t3p_PVw")
```

# Differential Analysis with DESeq2

## Prepare Object needed by DESeq2 for Analysis

Create two data.frame:

-   count_data : contain the raw counts of each genes for both control and PDCL samples
-   count_metadata : metadata for associated with the samples to know which samples belong to the controls and which belong to the PDCL

```{r prepare-count-data}
# Merge control and PDCL counts into one dataframe
count_data <- merge(control, pdcl_count_table, by.x = "symbol", by.y = "gene_id")
# Associate each row with a gene symbol
row.names(count_data) <- count_data$symbol
# Remove symbol and id columns
count_data <- count_data %>% dplyr::select(!c(symbol, id))
condition_vect <- c(
  rep("control", length(control_names)), 
  rep("pdcl", length(pdcl_names))
)
sample <- gsub("^(.*)_(Br\\d.*)", "\\1", colnames(count_data), ignore.case = T, perl = T, fixed = F)
# sample[(length(control_names)+1):length(sample)] <- "gbm"
replicate <- gsub("^(.*)_(Br\\d.*)", "\\2", colnames(count_data), ignore.case = T, perl = T, fixed = F)
count_metadata <- data.frame(
  sample = as.factor(sample),
  replicate = as.factor(replicate),
  condition = as.factor(condition_vect),
  row.names = colnames(count_data)
)
```

## Perform Differential Analysis

Perform a differential analysis for each PDCL in the database against the controls 
and save the result in rds file and a csv file.

```{r run-deseq2, cache=TRUE}
alpha <- 0.05
dds <- DESeqDataSetFromMatrix(
  countData = count_data, 
  colData = count_metadata, 
  design = ~condition
)
dds <- DESeq(dds)
deseq2_res <- results(dds, alpha = alpha, contrast = c("condition", "pdcl", "control"))

saveRDS(deseq2_res, file = paste0(result_dir, "deseq2_pdcl_vs_control.rds"))
write.csv(as.data.frame(deseq2_res), file = paste0(result_dir, "deseq2/deseq2_1_patient.csv"))

deregulated_index <- which(deseq2_res$padj <= alpha)
deregulated_genes <- row.names(deseq2_res)[deregulated_index]

summary(deseq2_res)
```

## VolcanoPlot Differential Analysis Results

```{r volcano-plot-deseq2, echo=FALSE}
deseq2_res_df <- as.data.frame(deseq2_res)
ggplot(
  deseq2_res_df, 
  aes(x = log2FoldChange, y = -log10(padj), colour = (padj<alpha) )
) + 
  geom_point() +
  scale_color_manual(values = c("TRUE" = "red"), labels = c("Deregulated")) +
  labs(
    title = "Volcano Plot DESeq2 Results",
    colour = NULL
  )
```

## Volcano Plot Super Deregulated genes
Number of super-deregulated genes (deregulated accross all the PDCL):

```{r super-deregulated-genes, echo=FALSE}
penda_genes <- row.names(penda_res)
pdcl_dereg_genes <- imap_dfr(penda_res, function(x, y){
  dereg_index <- which(x != 0)
  data.frame(gene = penda_genes[dereg_index], dereg = x[dereg_index], pdcl = y)
})

sd_dereg_genes <- pdcl_dereg_genes %>%
  count(gene, dereg) %>%
  filter(n==length(pdcl_names)) %>%
  arrange(gene) %>% 
  select(gene, dereg)

table(sd_dereg_genes$dereg)
```

```{r plot-sd-genes-lfc}
deseq2_res_df$sd_dereg <- 0
sd_index <- which(row.names(deseq2_res_df) %in% sd_dereg_genes$gene)
keep <- row.names(deseq2_res_df)[sd_index]
deseq2_res_df$sd_dereg[sd_index] <- sd_dereg_genes$dereg[sd_dereg_genes$gene %in% keep]
deseq2_res_df$sd_dereg <- as.factor(deseq2_res_df$sd_dereg)
data <- deseq2_res_df # %>% filter(sd_dereg != "0")
data$gene <- row.names(data)
volcano_sd_genes_lfc <- ggplot(
  data,
  aes(x = log2FoldChange, y = -log10(padj), colour = sd_dereg, alpha = 0.6, label = gene)
) + 
  geom_point() +
  scale_color_manual(values = c("1" = "red", "-1" = "blue"), labels = c("Super Up", "Super Down")) +
  labs(
    title = "Volcano Plot DESeq2 Results",
    colour = NULL
  )

volcano_sd_genes_lfc
```

## Compare DESeq2_log2FC with manually computed log2FC

```{r compare-lfc}
pdcl_mean_expr <- rowMeans(pdcl_count_table[,-1])
names(pdcl_mean_expr) <- pdcl_count_table$gene_id
control_mean_expr <- rowMeans(control[,c(-1,-2)])
names(control_mean_expr) <- control$symbol
pdcl_genes <- pdcl_count_table$gene_id[ pdcl_count_table$gene_id %in% control$symbol]
pdcl_mean_expr <- pdcl_mean_expr[pdcl_genes]
control_mean_expr <- control_mean_expr[pdcl_genes]
log2FC <- log2(pdcl_mean_expr/control_mean_expr)
names(log2FC) <- pdcl_genes

deseq2_FC <- deseq2_res$log2FoldChange
names(deseq2_FC) <- row.names(deseq2_res)
deseq2_FC <- deseq2_FC[pdcl_genes]

data <- data.frame(log2FC = log2FC, deseq2_FC = deseq2_FC)
data <- data %>% filter(!is.infinite(log2FC), !is.nan(log2FC), !is.infinite(deseq2_FC), !is.nan(deseq2_FC) )
fit <- lm(formula = deseq2_FC ~ log2FC, data = data)

p_comparison_log2fc <- ggplot(
  data,
  aes(x = log2FC, y = deseq2_FC)
) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  geom_abline(slope = fit$coefficients[2], intercept = fit$coefficients[1], color = "blue") +
  labs(title = "Comparison log2FoldChange", x = "log2FC manually computed", y = "log2FC computed by DESeq2")

p_comparison_log2fc
```

# Pathways Enrichment

```{r create-all-res-list}
all_res <- list()
```


## GSEA Pathway Enrichment

Here we run a GSEA pathway enrichment analysis with the log2FoldChange calculated by DESeq2 as the scoring metrics in the ranked list submited to GSEA.

*Message from Jorge :*

We must choose a seed to ensure the analysis is reproducible. This is because GSEA uses a permutation test to span the null distribution of the statistic. Hence, each time we run it, results are expected to change slightly.

```{r run-gsea, cache=TRUE}
# Create the ranked list for GSEA
# TODO: Check but it seems that using the stat of all genes after dropping the
# NAs seems to give the best result for GSEA (at least there is common 
# pathways between gprofiler and gsea)
rnk_tmp <- as.data.frame(deseq2_res) %>%
  drop_na() %>%
  # filter(padj < 0.1) %>%
  # mutate(rank = log2FoldChange*-log10(padj)) %>%
  mutate(rank = stat) %>%
  arrange(desc(rank))
rnk <- rnk_tmp %>% pull(rank)
names(rnk) <- row.names(rnk_tmp)
rnk <- rnk[is.finite(rnk)]
rank_list <- list("gbm" = rnk)

gmt_desc <- lapply(gmt_path, function(x){
  gmt_info <- read.gmt(x)
  lapply(gmt_info, function(x){ x$description })
})

# Run GSEA analysis for each ranked list for both Reactome and KEGG
all_res[["GSEA"]] <- run.gsea(rank_list, gsea_gmt_list, gmt_desc)
```

## GProfiler Pathway Enrichment

Run G:Profiler

```{r run-gprofiler, cache=TRUE}
all_pdcl_genes <- row.names(deseq2_res)
all_gene_list <- list("gbm" = deregulated_genes)

# TODO: Run G:Profiler analysis for each db and samples
all_res[["G:Profiler"]] <- run.gost(all_gene_list, gprofiler_gmt_list, all_pdcl_genes)
```

# Save Result in RDS and in Generic Enrichment Map (GEM) format

```{r save-rds}
saveRDS(all_res, file = "Result_pdcl_process_as_1_patient/all_res.rds")
```

```{r convert-to-enrichment}
all_enrichment <- list()

# TODO: Add pathway_desc to enrichment mapping
all_enrichment[["GSEA"]] <- map_depth(all_res[["GSEA"]], .depth = 1, .f = convert.gsea.to.enrichment)

all_enrichment[["G:Profiler"]] <- map_depth(all_res[["G:Profiler"]], .depth = 1, .f = convert.gost.to.enrichment)
```

```{r save-enrichment}
save.gsea.gem.list.to.txt(all_enrichment[["GSEA"]], paste0(result_dir, "gsea/"))
save.gost.gem.list.to.txt(all_enrichment[["G:Profiler"]], paste0(result_dir, "gprofiler/"))
```

## Process Result

### Collapse Enrichment

```{r collapse-by-method}
# Collapse the results by method.
# One line contain the result of one enrichment method for one pathway of one PDCL
# TODO! Redefine how it will be collapsed
all_enrich_collapsed <- collapse.enrichment(all_enrichment)
threshold <- 0.1
```

```{r count-enriched-pathways}
data <- all_enrich_collapsed %>% filter(fdr<threshold)
count_enrich_pathways <- ggplot(data, aes(x = method, fill = db) ) +
  geom_bar(position = "dodge2")
count_enrich_pathways
```

Number of commons pathways per samples and databases :
```{r check-number-of-commons pathways}
all_enrich_collapsed %>% filter(fdr < threshold) %>% count(sample, db, pathway_id) %>% filter(n==2) %>% count(sample, db)
```

```{r find-common-pathways}
common_pathway_id <- all_enrich_collapsed %>% 
  filter(fdr < threshold) %>% 
  dplyr::count(sample, db, pathway_id) %>% 
  filter(n==2) %>% 
  pull(pathway_id)
```

```{r download-go-chart}
id2go <- read.table("Data/gene-set/id2go.txt", sep = "\t", header = T)
common_go_ids <- id2go %>% dplyr::filter(source_id %in% common_pathway_id) %>% pull(go_id)
download.go.chart(common_go_ids, paste0(result_dir, "plot/alpha_", threshold, "/common_pathways_go_chart.png"))
```


```{r print-mapped-ids, echo=FALSE}
common_pathway_2_go <- all_enrich_collapsed %>% 
  dplyr::filter(pathway_id %in% common_pathway_id) %>% 
  dplyr::distinct(pathway_id, description, db) %>%
  dplyr::left_join(id2go, by = c("pathway_id" = "source_id")) %>%
  dplyr::arrange(go_id) %>%
  dplyr::select(pathway_id, description, db, go_id, go_desc)

DT::datatable(
  common_pathway_2_go,
  rownames = F,
  caption = "Commons pathways with their mapped GO Terms"
)
```

```{r save-session, show=FALSE}
save.image("~/PhD/PDCL/pdcl_process_as_1_patient.RData")
```

