---
title: "PDCL_process_personalized"
output: 
  flexdashboard::flex_dashboard:
    orientation: column
    vertical_layout: fill
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)
knitr::opts_chunk$set(cache=TRUE)

# Load required libraries (others are load directly in sourced files)
library(DESeq2)
library(dplyr)
library(purrr)
library(plotly)

# Source files with functions used in next steps
source("R/gprofiler-analysis.R")
source("R/gsea-analysis.R")
source("R/process-result.R")
source("R/quickgo.R")
source("R/kegg.R")
source("R/reactome.R")

# Constant
result_dir <- "~/PhD/PDCL/Result_pdcl_process_personnalized/"
```

```{r function}
# This function filter and prepare the data to be plotted in a fdr heatmap
prepare.fdr.data <- function(pattern){
  data <- all_enrich_collapsed %>%
    dplyr::inner_join( rbind(kegg_gs_metadata, react_gs_metadata) ) %>%
    dplyr::filter(
      pathway_id %in% common_pathways$pathway_id, 
      str_detect(category, regex(pattern, ignore_case = T))
    ) %>%
    dplyr::arrange(desc(description))
  
  data_commons <- common_pathways %>%
    dplyr::filter(pathway_id %in% data$pathway_id)
  
  ordered_description <- data %>% 
    dplyr::distinct(description, category) %>%
    dplyr::arrange(desc(description)) %>%
    dplyr::pull(description)
  
  ordered_category <- data %>% 
    dplyr::distinct(description, category) %>%
    dplyr::arrange(desc(description)) %>%
    dplyr::pull(category)
  
  data$description <- factor(data$description, levels = ordered_description, ordered = T)
  
  list(
    data = data, 
    data_commons = data_commons,
    ordered_description = ordered_description,
    ordered_category = ordered_category
  )
}
```

# Data Loading

Load controls

```{r load-controls, echo = TRUE}
control_file <- "Data/PDCL/astrocyte_GSE109001_counts_unique_genes_samples_filtered.txt"
control <- read.table(
  control_file, 
  header = T, 
  sep = "\t", 
  as.is = c("id", "symbol")
)
control_names <- c(
  "AF22_NES_Astro_Br1_d29_37_S46",
  "AF22_NES_Astro_Br2_d29_38_S56",
  "AF22_NES_Astro_Br3_d29_39_S66",
  "CCF.STTG1_p24_Br1_S16",
  "CCF.STTG1_p24_Br2_S17",
  "CCF.STTG1_p24_Br3_S18",
  "CDIAstrocytes_p2_Br1_S19",
  "CDIAstrocytes_p2_Br2_S20",
  "CDIAstrocytes_p2_Br3_S21",
  "phaAstrocyte_p2_Br1_S1",
  "phaAstrocyte_p2_Br2_S2",
  "phaAstrocyte_p2_Br3_S3"
  )
```

Load PDCL DataBase

Load the PDCL count table. Disabled check.names otherwise it will add a X at the beginning of each column name, as column names aren't valid according to R.

```{r load-pdcl, echo = TRUE}
pdcl_count_table <- read.table(
  "/home/spinicck/PhD/Data/PDCL/lignees_count_genes_PDCL.txt", 
  sep = "\t", 
  header = T, 
  as.is = "gene_id", 
  check.names = F
)

pdcl_names <- c(
  "4339-p21",
  "4371-p37",
  "5706-p14",
  "6190-p43",
  "6240-p12",
  "7015-p17",
  "7060-p18", 
  "7142-p14",
  "N13-1300",
  "N13-1520-p9",
  "N14-0072",
  "N14-0870",
  "N14-1208",
  "N14-1525",
  "N15_0460",
  "N15_0516",
  "N15-0385",
  "N15-0661",
  "N16_0535",
  "N16-0240"
)
```

Load PENDA results

```{r load-penda, echo = TRUE}
penda_res <- readRDS("Data/PDCL/results_penda_astrocytes_2021.rds")
penda_res <- as.data.frame(penda_res)
penda_pdcl_names <- names(penda_res)
# Remove ".genes.results" string in the column names
penda_pdcl_names <- sub(".genes.results", "", penda_pdcl_names, ignore.case = T)
names(penda_res) <- penda_pdcl_names
penda_genes <- row.names(penda_res)
```

Define GMT path

```{r define-gmt-path, echo = TRUE}
gmt_path <- list(
  reactome = "Data/gene-set/reactome_no_less_15_no_more_500.gmt",
  kegg = "Data/gene-set/kegg_no_less_15_no_more_500.gmt"
)
```

Load GMT files for GSEA pathway enrichment.

```{r create-gsea-gmt-list, echo = TRUE}
gsea_gmt_list <- lapply(gmt_path, function(x){ gmtPathways(x)})
```

Load G:Profiler GMTs

Upload GMT File before running G:Profiler Enrichment Analysis if needed. Otherwise juste define the token of the GMT stored on G:Profiler website.

```{r create-gprofiler-gmt-list, echo = TRUE}
# This portion is commented because the file are already uploaded
gprofiler_gmt_list <- lapply(gmt_path, function(x){ upload_GMT_file(x) })
# bioplanet_gmt_token <- "gp__vB7V_DT57_yNY"
# gprofiler_gmt_list <- list(reactome = "gp__Nqmx_yxm7_37U", kegg = "gp__5f4Q_8t3p_PVw")
# gprofiler_gmt_list <- list(reactome = "gp__VxvL_Ua0o_Irk", kegg = "gp__5f4Q_8t3p_PVw")
```

# DE Analysis

```{r prepare-count-data}
# Merge control and PDCL counts into one dataframe
count_data <- merge(control, pdcl_count_table, by.x = "symbol", by.y = "gene_id")
# Associate each row with a gene symbol
row.names(count_data) <- count_data$symbol
# Remove symbol and id columns
count_data <- count_data %>% dplyr::select(!c(symbol, id))
condition_vect <- c(
  rep("control", length(control_names)), 
  rep("pdcl", length(pdcl_names))
)
count_metadata <- data.frame(
  sample = c(control_names, pdcl_names), 
  condition=condition_vect
)
count_metadata$condition <- as.factor(count_metadata$condition)
```

```{r run-deseq2, cache=TRUE}
deseq2_res_list <- lapply(pdcl_names, function(pdcl){
  dataset_pdcl <- c(control_names, pdcl)
  # Subract a dataset which contains only the counts for the
  # current PDCL
  dataset <- count_data[,dataset_pdcl]
  # Same with metadata
  dataset_metadata <- count_metadata[ count_metadata$sample %in% dataset_pdcl, ]
  # Perform DESeq2 Analysis
  dds <- DESeqDataSetFromMatrix(
    countData = dataset, 
    colData = dataset_metadata, 
    design = ~condition
  )
  dds <- DESeq(dds)
  res <- results(dds, alpha = 0.05)
})
names(deseq2_res_list) <- pdcl_names
```


```{r save-deseq2}
# Save Result in rds file and write a csv file for each PDCL.
saveRDS(deseq2_res_list, file = paste0(result_dir, "deseq2_pdcl_vs_control.rds"))

for (pdcl in pdcl_names){
  deseq2_res <- deseq2_res_list[[pdcl]]
  res.file <-  paste0(result_dir, "deseq2/", pdcl, ".csv")
  write.csv(as.data.frame(deseq2_res), file = res.file)
}
```


## Column {.tabset}

### Volcano -log10(padj)

```{r collapse-de-result}
de_res <- lapply(deseq2_res_list, function(x){
  df <- as.data.frame(x)
  df$gene <- row.names(x)
  row.names(df) <- NULL
  df
})
de_res <- map.list.to.df(de_res, "sample")

penda_res_collapsed <- imap_dfr(penda_res, function(x, y){
  data.frame(gene = penda_genes, penda_predict = x, sample = y)
})

de_res <- de_res %>%
  dplyr::inner_join(penda_res_collapsed) %>%
  dplyr::mutate(penda_predict = as.factor(penda_predict)) %>%
  dplyr::relocate(sample, gene)
```


```{r volcano-plot-log10padj, fig.width=14, fig.height=8}
volcano_plot_log10padj <- ggplot(
    de_res, 
    aes( log2FoldChange, -log10(padj), colour = penda_predict, alpha = penda_predict )
  ) +
    geom_point() +
    scale_colour_manual( values = c("1" = "red", "-1"="blue", "0" = "grey50") )  +
    scale_alpha_manual(values = c("1" = 0.8, "-1" = 0.8, "0" = 0.4) ) +
    labs(
      title = "Comparison DESeq2 fold change vs PENDA Result",
      colour = "Penda Prediction", 
      alpha = "Penda Prediction"
    ) + 
  facet_wrap(~sample)
volcano_plot_log10padj
```


### Volcano abs(stat)

```{r volcano-plot-stat, fig.width=14, fig.height=8}
volcano_plot_stat <- ggplot(
    de_res, 
    aes( log2FoldChange, abs(stat), colour = penda_predict, alpha = penda_predict )
  ) +
    geom_point() +
    scale_colour_manual( values = c("1" = "red", "-1"="blue", "0" = "grey50") )  +
    scale_alpha_manual(values = c("1" = 0.8, "-1" = 0.8, "0" = 0.4) ) +
    labs(
      title = "Comparison DESeq2 fold change vs PENDA Result",
      colour = "Penda Prediction", 
      alpha = "Penda Prediction"
    ) + 
  facet_wrap(~sample)
volcano_plot_stat
```


```{r create-all-res-list}
all_res <- list()
```

```{r run-gsea, cache=TRUE}
set.seed(1)

rank_list <- lapply(deseq2_res_list, function(x){
  rnk <- x$stat
  names(rnk) <- row.names(x)
  rnk <- rnk[!is.na(rnk) & is.finite(rnk)]
  rnk <- sort(rnk, decreasing = T)
  rnk
})

gmt_desc <- lapply(gmt_path, function(x){
  gmt_info <- read.gmt(x)
  lapply(gmt_info, function(x){ x$description })
})

# Run GSEA analysis for each ranked list for both Reactome and KEGG
all_res[["GSEA"]] <- run.gsea(rank_list, gsea_gmt_list, gmt_desc)
```

```{r run-gprofiler, cache=TRUE}
all_pdcl_genes <- row.names(penda_res)
all_gene_list <- lapply(penda_res, function(x){
  all_pdcl_genes[x != 0]
})
all_res[["G:Profiler"]] <- run.gost(all_gene_list, gprofiler_gmt_list, all_pdcl_genes)
```

```{r convert-to-enrichment}
all_enrichment <- list()

# TODO: Add pathway_desc to enrichment mapping
all_enrichment[["GSEA"]] <- map_depth(all_res[["GSEA"]], .depth = 1, .f = convert.gsea.to.enrichment)

all_enrichment[["G:Profiler"]] <- map_depth(all_res[["G:Profiler"]], .depth = 1, .f = convert.gost.to.enrichment)
```

```{r collapse-by-method}
# Collapse the results by method.
# One line contain the result of one enrichment method for one pathway of one PDCL
# TODO! Redefine how it will be collapsed
all_enrich_collapsed <- collapse.enrichment(all_enrichment)
threshold <- 0.1
```

```{r find-common-pathways}
common_pathways <- all_enrich_collapsed %>% 
  filter(fdr < threshold) %>% 
  dplyr::count(sample, db, pathway_id, description) %>% 
  filter(n==2)
```

```{r gprofiler-vs-gsea}
gprofiler_res <- all_enrich_collapsed %>%
    filter(method == "G:Profiler")
  gsea_res <- all_enrich_collapsed %>%
    filter(method == "GSEA")
  gprofiler_vs_gsea <- gprofiler_res %>%
    dplyr::full_join(gsea_res, by = c("pathway_id", "description", "db", "sample")) %>%
    dplyr::rename(pval_gprofiler = p_value.x, pval_gsea = p_value.y, fdr_gprofiler = fdr.x, fdr_gsea = fdr.y) %>%
    select(sample, db, pathway_id, description, pval_gprofiler, pval_gsea, fdr_gprofiler, fdr_gsea)
```

# Kegg {data-navmenu="Comparison Enrichment"}

### Comparison Enrichment

```{r compare-enrichment-kegg, fig.width=14, fig.height=8}
data <- gprofiler_vs_gsea %>%
  dplyr::filter(db == "kegg")
  
comparison_kegg <- ggplot(
  data = data, 
  aes(
    x = -log10(fdr_gprofiler), 
    y = -log10(fdr_gsea),
    colour = (fdr_gprofiler<threshold & fdr_gsea<threshold)
  )
) +
  geom_point(
    size = 0.5
  ) +
  scale_colour_manual(values = c("TRUE" = "red", "FALSE" = "grey30"), label = c("Common", "Specific")) +
  facet_wrap(vars(sample)) +
  labs(
    title = "Comparison fdr of G:Profiler vs GSEA for the Kegg Database",
    subtitle = "Threshold alpha : 0.1",
    x = paste0( "-log10(fdr G:Profiler)"),
    y = paste0( "-log10(fdr GSEA)"),
    colour = ""
  ) +
  theme(
    plot.title = element_text(hjust = 0.5,face = "bold",color = "red"),
    plot.subtitle = element_text(hjust = 0.5,face = "italic",color = "blue"),
    axis.text.x = element_text(angle = 90)
  ) 
comparison_kegg
```

# Reactome {data-navmenu="Comparison Enrichment"}

### Comparison Enrichment

```{r compare-enrichment-reactome, fig.width=14, fig.height=8}
data <- gprofiler_vs_gsea %>%
  dplyr::filter(db == "reactome")
  
comparison_reactome <- ggplot(
  data = data, 
  aes(
    x = -log10(fdr_gprofiler), 
    y = -log10(fdr_gsea),
    colour = (fdr_gprofiler<threshold & fdr_gsea<threshold)
  )
) +
  geom_point(
    size = 0.5
  ) +
  scale_colour_manual(values = c("TRUE" = "red", "FALSE" = "grey30"), label = c("Common", "Specific")) +
  facet_wrap(vars(sample)) +
  labs(
    title = "Comparison fdr of G:Profiler vs GSEA  for the Reactome Database",
    subtitle = "Threshold alpha : 0.1",
    x = paste0( "-log10(fdr G:Profiler)"),
    y = paste0( "-log10(fdr GSEA)"),
    colour = ""
  ) +
  theme(
    plot.title = element_text(hjust = 0.5,face = "bold",color = "red"),
    plot.subtitle = element_text(hjust = 0.5,face = "italic",color = "blue"),
    axis.text.x = element_text(angle = 90)
  ) 
comparison_reactome
```


<!-- Retrieve Categories for enriched pathways -->

```{r kegg-categories, cache=TRUE}
kegg_gs_metadata <- read.csv("Data/gene-set/kegg_geneset_metadata.csv", as.is = T)
kegg_categories <- factor(kegg_gs_metadata$category, levels = KEGG_TOP_CATEGORIES)
names(kegg_categories) <- kegg_gs_metadata$pathway_id
```

```{r reactome-categories, cache=TRUE}
react_gs_metadata <- read.csv("Data/gene-set/reactome_geneset_metadata.csv", as.is = T)
react_categories <- factor(react_gs_metadata$category, levels = REACTOME_TOP_LEVEL_PATHWAYS)
names(react_categories) <- react_gs_metadata$pathway_id
```

# Kegg {data-navmenu="Category Enriched"}

### Nb Pathway Enriched per Categories

```{r nb-pathway-enriched-per-kegg-categories, fig.width=16, fig.height=10}
count <- all_enrich_collapsed %>%
  dplyr::filter(fdr < threshold, db == "kegg") %>%
  dplyr::mutate(category = kegg_categories[pathway_id] ) %>%
  dplyr::count(sample, method, category, .drop = F)

enriched_categories_kegg <- ggplot(
  data = count,
  aes(x = sample, y = category, fill = n)
) + 
  geom_raster() +
  labs(
    title = "Number of pathways enriched in each categories for each PDCL",
    x = "PDCL",
    y = "Categories",
    fill = "Count"
  ) +
  facet_wrap(~method) +
  scale_fill_viridis_c(direction = 1) +
  theme(axis.text.x = element_text(angle = 90) ) +
  # By default it is sorted in desc order, we reverse it so it is in asc order
  scale_y_discrete(limits=rev)
enriched_categories_kegg
```

# Reactome {data-navmenu="Category Enriched"}

### Nb Pathway Enriched per Categories

```{r nb-pathway-enriched-per-reactome-categories, fig.width=16, fig.height=10}
count <- all_enrich_collapsed %>%
  dplyr::filter(fdr < threshold, db == "reactome") %>%
  dplyr::mutate(category = react_categories[pathway_id] ) %>%
  dplyr::count(sample, method, category, .drop = F)

enriched_categories_reactome <- ggplot(
  data = count,
  aes(x = sample, y = category, fill = n)
) + 
  geom_raster() +
  labs(
    title = "Number of pathways enriched in each categories for each PDCL",
    x = "PDCL",
    y = "Categories",
    fill = "Count"
  ) +
  facet_wrap(~method) +
  scale_fill_viridis_c(direction = 1) +
  theme(axis.text.x = element_text(angle = 90) ) +
  # By default it is sorted in desc order, we reverse it so it is in asc order
  scale_y_discrete(limits=rev)
enriched_categories_reactome
```

# Commons pathways {data-navmenu="Category Enriched" data-orientation=rows}

### Kegg

```{r common-pathways-kegg, fig.width=14, fig.height=8}
kegg_commons <- common_pathways %>%
  dplyr::inner_join( rbind(kegg_gs_metadata, react_gs_metadata)) %>%
  dplyr::filter(db == "kegg") %>%
  dplyr::select(!c(n, size))

barplot_kegg_commons <- ggplot(kegg_commons, aes(x = sample, fill = category)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90) )
barplot_kegg_commons
```


### Reactome

```{r common-pathways-reactome, fig.width=14, fig.height=8}
reactome_commons <- common_pathways %>%
  dplyr::inner_join( rbind(kegg_gs_metadata, react_gs_metadata)) %>%
  dplyr::filter(db == "reactome") %>%
  dplyr::select(!c(n, size))

barplot_reactome_commons <- ggplot(reactome_commons, aes(x = sample, fill = category)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90) )
barplot_reactome_commons
```

# Count Enriched Pathways {data-orientation=rows}

### Count Enriched pathways per method

```{r count-enriched-method, fig.width=14, fig.height=8}
data <- all_enrich_collapsed %>%
  dplyr::filter(fdr<threshold)
count_enriched_method <- ggplot(
  data = data, 
  aes(x = sample, fill = db)
) +
  geom_bar(position = "dodge2") +
  labs(
    title = "Number of enriched pathways for each enrichment method",
    x = "Method",
    y = "Counts",
    fill = ""
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~method, nrow = 2, ncol = 1) +
  theme(axis.text.x = element_text(angle = 90) )
count_enriched_method
```

### Count commons per PDCL

```{r count-common-pdcl, fig.width=14, fig.height=8}
count_common <- ggplot(
  data = common_pathways, 
  aes(x = sample, fill = db) 
) +
  geom_bar(position = "dodge2") +
  labs(
    title = "Number of common pathways for each pdcl",
    x = "Method",
    y = "Counts",
    fill = ""
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  theme(axis.text.x = element_text(angle = 90) )
count_common
```


# Enrichment Per PDCL

## Column {.tabset}

### Heatmap FDR Metabolism and ECM

```{r heatmap-fdr-metabolism-ecm, fig.width=18, fig.height=12}
heatmap_fdr_metabolism_ecm <- with( prepare.fdr.data("\\bmetabolism|matrix\\b") ,{
  heatmap.fdr(data, data_commons) +
  guides(
    y.sec =  guide_axis_manual(
      labels = ordered_category,
      title = "Category"
    )
  )
})
heatmap_fdr_metabolism_ecm
```

### Heatmap FDR Cell Cycle

```{r heatmap-fdr-cell-cycle-and-signalling, fig.width=18, fig.height=12}
heatmap_fdr_cell_cycle_signal <- with( prepare.fdr.data("\\b(cell.*|dna|signal|processing)\\b") ,{
  heatmap.fdr(data, data_commons) +
  guides(
    y.sec =  guide_axis_manual(
      labels = ordered_category,
      title = "Category"
    )
  )
})
heatmap_fdr_cell_cycle_signal
```

```{r save-session, show=FALSE}
save.image("~/PhD/PDCL/pdcl_process_personalized.RData")
```
