---
title: "1 - PDCL Enrichment"
output: 
  flexdashboard::flex_dashboard:
    orientation: column
    vertical_layout: fill
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)

# Load required libraries (others are load directly in sourced files)
library(DESeq2)
library(gprofiler2)
library(dplyr)
library(purrr)
library(plotly)
library(Biobase)

# Source files with functions used in next steps
source("../Code/R/pdcl-dataset.R")
source("../Code/R/gprofiler-analysis.R")
source("../Code/R/gsea-analysis.R")
source("../Code/R/process-result.R")
source("../Code/R/quickgo.R")
source("../Code/R/kegg.R")
source("../Code/R/reactome.R")

# Constant
rerun_rds_cached <- T
result_dir <- "Result_1_pdcl_enrichment/"
```

# Data Loading

Define GMT path

```{r define-gmt-path, echo = TRUE}
gmt_path <- list(
  reactome = "~/PhD/Project/PDCL/Data/gene-set/reactome_no_less_15_no_more_500.gmt",
  kegg = "~/PhD/Project/PDCL/Data/gene-set/kegg_no_less_15_no_more_500.gmt"
)
```

Load GMT files for GSEA pathway enrichment.

```{r create-gsea-gmt-list, echo = TRUE}
gsea_gmt_list <- lapply(gmt_path, function(x){ gmtPathways(x)})
```

Load G:Profiler GMTs

Upload GMT File before running G:Profiler Enrichment Analysis if needed. Otherwise juste define the token of the GMT stored on G:Profiler website.

```{r create-gprofiler-gmt-list, echo = TRUE}
# This portion is commented because the file are already uploaded
gprofiler_gmt_list <- lapply(gmt_path, function(x){ upload_GMT_file(x) })
# bioplanet_gmt_token <- "gp__vB7V_DT57_yNY"
# gprofiler_gmt_list <- list(reactome = "gp__Nqmx_yxm7_37U", kegg = "gp__5f4Q_8t3p_PVw")
# gprofiler_gmt_list <- list(reactome = "gp__VxvL_Ua0o_Irk", kegg = "gp__5f4Q_8t3p_PVw")
```

# DESeq2

```{r prepare-count-data}
control_count_data <- read.csv(PDCL_DATASET_PARAM$control_count_path)
pdcl_count_data <- read.csv(PDCL_DATASET_PARAM$pdcl_count_path, check.names = F)
# Merge control and PDCL counts into one dataframe
count_data <- merge(control_count_data, pdcl_count_data, by.x = "symbol", by.y = "gene_id")
# Associate each row with a gene symbol
row.names(count_data) <- count_data$symbol
# Remove symbol and id columns
count_data <- count_data %>% dplyr::select(!c(symbol, id))
condition_vect <- c(
  rep("control", length(PDCL_DATASET_PARAM$control_sample_names)), 
  rep("pdcl", length(PDCL_DATASET_PARAM$pdcl_sample_names))
)
sample <- gsub("^(.*)_(Br\\d.*)", "\\1", colnames(count_data), ignore.case = T, perl = T, fixed = F)
# sample[(length(control_names)+1):length(sample)] <- "gbm"
replicate <- gsub("^(.*)_(Br\\d.*)", "\\2", colnames(count_data), ignore.case = T, perl = T, fixed = F)
count_metadata <- data.frame(
  sample = as.factor(sample),
  replicate = as.factor(replicate),
  condition = as.factor(condition_vect),
  row.names = colnames(count_data)
)
```

## Column {data-width="400"}

### DESeq2 Result {data-height="300"}

```{r run-deseq2}
alpha <- 0.05
dds <- DESeqDataSetFromMatrix(
  countData = count_data, 
  colData = count_metadata, 
  design = ~condition
)
dds <- DESeq(dds)

deseq2_res <- xfun::cache_rds({
  results(dds, alpha = alpha, contrast = c("condition", "pdcl", "control"))
}, file = "deseq2_pdcl_vs_control.rds", dir = result_dir, rerun = rerun_rds_cached)

write.csv(as.data.frame(deseq2_res), file = paste0(result_dir, "deseq2/deseq2_1_patient.csv"))

deregulated_index <- which(deseq2_res$padj <= alpha)
deregulated_genes <- row.names(deseq2_res)[deregulated_index]

summary(deseq2_res)
```

### Comparison Log2FoldChange {data-height="700"}

```{r compare-lfc}
pdcl_mean_expr <- rowMeans(pdcl_count_data[,-1])
names(pdcl_mean_expr) <- pdcl_count_data$gene_id
control_mean_expr <- rowMeans(control_count_data[,c(-1,-2)])
names(control_mean_expr) <- control_count_data$symbol
pdcl_genes <- pdcl_count_data$gene_id[ pdcl_count_data$gene_id %in% control_count_data$symbol]
pdcl_mean_expr <- pdcl_mean_expr[pdcl_genes]
control_mean_expr <- control_mean_expr[pdcl_genes]
log2FC <- log2(pdcl_mean_expr/control_mean_expr)
names(log2FC) <- pdcl_genes

deseq2_FC <- deseq2_res$log2FoldChange
names(deseq2_FC) <- row.names(deseq2_res)
deseq2_FC <- deseq2_FC[pdcl_genes]

data <- data.frame(log2FC = log2FC, deseq2_FC = deseq2_FC)
data <- data %>% filter(!is.infinite(log2FC), !is.nan(log2FC), !is.infinite(deseq2_FC), !is.nan(deseq2_FC) )
fit <- lm(formula = deseq2_FC ~ log2FC, data = data)

p_comparison_log2fc <- ggplot(
  data,
  aes(x = log2FC, y = deseq2_FC)
) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  geom_abline(slope = fit$coefficients[2], intercept = fit$coefficients[1], color = "blue") +
  labs(title = "Comparison log2FoldChange", x = "log2FC manually computed", y = "log2FC computed by DESeq2")

p_comparison_log2fc
```

## Column {data-width="600" .tabset}

### Volcano Plot

```{r volcano-plot-deseq2, fig.width=12, fig.height=8}
deseq2_res_df <- as.data.frame(deseq2_res)
ggplot(
  deseq2_res_df, 
  aes(x = log2FoldChange, y = -log10(padj), colour = (padj<alpha) )
) + 
  geom_point() +
  scale_color_manual(values = c("TRUE" = "red"), labels = c("Deregulated")) +
  labs(
    title = "Volcano Plot DESeq2 Results",
    colour = NULL
  )
```

### Volcano Plot Stat DESeq2

```{r volcano-stat, fig.width=12, fig.height=8}
ggplot(
  deseq2_res_df, 
  aes(x = log2FoldChange, y = -log10(padj), colour = stat )
) + 
  geom_point() +
  scale_colour_distiller(palette = "RdBu", n.breaks = 8) +
  labs(
    title = "Volcano Plot DESeq2 Results",
    colour = "DESeq2 Statistic"
  ) +
  theme_dark()
```

```{r create-all-res-list}
all_res <- list()
```

```{r run-gsea}
# Create the ranked list for GSEA
# TODO: Check but it seems that using the stat of all genes after dropping the
# NAs seems to give the best result for GSEA (at least there is common 
# pathways between gprofiler and gsea)
rnk_tmp <- as.data.frame(deseq2_res) %>%
  drop_na() %>%
  # filter(padj < 0.1) %>%
  # mutate(rank = log2FoldChange*-log10(padj)) %>%
  mutate(rank = stat) %>%
  arrange(desc(rank))
rnk <- rnk_tmp %>% pull(rank)
names(rnk) <- row.names(rnk_tmp)
rnk <- rnk[is.finite(rnk)]
rank_list <- list("gbm" = rnk)

gmt_desc <- lapply(gmt_path, function(x){
  gmt_info <- read.gmt(x)
  lapply(gmt_info, function(x){ x$description })
})

# Run GSEA analysis for each ranked list for both Reactome and KEGG
all_res[["GSEA"]] <- xfun::cache_rds({
  all_res[["GSEA"]] <- run.gsea(rank_list, gsea_gmt_list, gmt_desc)
}, file = "gsea_results.rds", dir = result_dir, rerun = rerun_rds_cached)
```

```{r run-gprofiler}
all_pdcl_genes <- row.names(deseq2_res)
all_gene_list <- list("gbm" = deregulated_genes)
all_res[["G:Profiler"]] <- xfun::cache_rds({
  all_res[["G:Profiler"]] <- run.gost(all_gene_list, gprofiler_gmt_list, all_pdcl_genes)
}, file = "gprofiler_results.rds", dir = result_dir, rerun = rerun_rds_cached)
```

```{r convert-to-enrichment}
all_enrichment <- list()

all_enrichment[["GSEA"]] <- map_depth(all_res[["GSEA"]], .depth = 1, .f = convert.gsea.to.enrichment)

all_enrichment[["G:Profiler"]] <- map_depth(all_res[["G:Profiler"]], .depth = 1, .f = convert.gost.to.enrichment)
```

```{r save-enrichment}
save.gsea.gem.list.to.txt(all_enrichment[["GSEA"]], paste0(result_dir, "gsea/"))
save.gost.gem.list.to.txt(all_enrichment[["G:Profiler"]], paste0(result_dir, "gprofiler/"))
```

```{r collapse-by-method}
# Collapse the results by method.
# One line contain the result of one enrichment method for one pathway of one PDCL
# TODO! Redefine how it will be collapsed
all_enrich_collapsed <- collapse.enrichment(all_enrichment)
threshold <- 0.1
```

```{r find-common-pathways}
common_pathway_id <- all_enrich_collapsed %>% 
  filter(fdr < threshold) %>% 
  dplyr::count(sample, db, pathway_id) %>% 
  filter(n==2) %>% 
  pull(pathway_id)
```

```{r download-go-chart}
id2go <- read.table("../Data/gene-set/id2go.txt", sep = "\t", header = T, as.is = T)
common_go_ids <- id2go %>% dplyr::filter(source_id %in% common_pathway_id) %>% pull(go_id)

# Download GO chart for common GO IDs
download.go.chart(common_go_ids, paste0(result_dir, "plot/common_pathways_go_chart.png"))
```

# Enrichment Score {data-navmenu="GSEA" data-orientation=rows}

### Reactome

```{r plot-nes-reactome, fig.width=20, fig.height=8}
data <- all_res[["GSEA"]]$reactome$gbm %>%
  dplyr::filter(pathway %in% common_pathway_id, padj < threshold) %>%
  dplyr::arrange(padj) %>%
  dplyr::slice(1:25)
nes_plot_reactome <- plot.nes.gsea(data)
nes_plot_reactome
```

### Kegg

```{r plot-nes-kegg, fig.width=9, fig.height=8}
data <- all_res[["GSEA"]]$kegg$gbm %>%
  dplyr::filter(pathway %in% common_pathway_id, padj < threshold) %>%
  dplyr::arrange(padj) %>%
  dplyr::slice(1:25)
nes_plot_kegg <- plot.nes.gsea(data)
nes_plot_kegg
```

# Leading Edge {data-navmenu="GSEA"}

### Reactome

```{r leading-edge-reactome, fig.width=20}
gsea_reactome_enriched <- all_enrich_collapsed %>%
  dplyr::filter(pathway_id %in% common_pathway_id, method == "GSEA", fdr < threshold, db == "reactome") %>%
  dplyr::arrange(fdr) %>%
  dplyr::slice(1:25)
gsea_reactome_leadedge <- plot.leading.edge(gsea_reactome_enriched, rank_list$gbm)
# ggplotly(gsea_reactome_leadedge)
gsea_reactome_leadedge
```

### Kegg

```{r leading-edge-kegg, fig.width=20}
gsea_kegg_enriched <- all_enrich_collapsed %>%
  dplyr::filter(pathway_id %in% common_pathway_id, method == "GSEA", fdr < threshold, db == "kegg") %>%
  dplyr::arrange(fdr) %>%
  dplyr::slice(1:25)
gsea_kegg_leadedge <- plot.leading.edge(gsea_kegg_enriched, rank_list$gbm)
# ggplotly(gsea_kegg_leadedge)
gsea_kegg_leadedge
```

# Reactome {data-navmenu="Count Pathways Per genes"}

### GSEA

```{r count-path-per-genes-reac-gsea, fig.width=20}
data <- all_enrich_collapsed %>% 
  dplyr::filter(fdr < threshold, db == "reactome", method == "GSEA")
reactome_count_path_per_genes <- plot.number.pathways.per.genes(data)
reactome_count_path_per_genes
```

### G:Profiler

```{r count-path-per-genes-react-gprofiler, fig.width=20}
data <- all_enrich_collapsed %>% 
  dplyr::filter(fdr < threshold, db == "reactome", method == "G:Profiler")
reactome_count_path_per_genes <- plot.number.pathways.per.genes(data)
reactome_count_path_per_genes
```

# Kegg {data-navmenu="Count Pathways Per genes"}

### GSEA

```{r count-path-per-genes-kegg-gsea, fig.width=20}
data <- all_enrich_collapsed %>% 
  dplyr::filter(fdr < threshold, db == "kegg", method == "GSEA")
kegg_count_path_per_genes <- plot.number.pathways.per.genes(data)
kegg_count_path_per_genes
```

### G:Profiler

```{r count-path-per-genes-kegg-gprofiler, fig.width=20}
data <- all_enrich_collapsed %>% 
  dplyr::filter(fdr < threshold, db == "kegg", method == "G:Profiler")
kegg_count_path_per_genes <- plot.number.pathways.per.genes(data)
# ggplotly(kegg_count_path_per_genes)
kegg_count_path_per_genes
```

# Enrichment and Categories

## Column {data-width=300}

### Number of Enriched Pathways

```{r count-enriched-pathways}
data <- all_enrich_collapsed %>% filter(fdr<threshold)
count_enrich_pathways <- plot.count.pathways(data)
count_enrich_pathways
```

## Column {data-width=700}

### Number of Enriched Kegg Categories
<!-- Retrieve Categories for enriched pathways -->

```{r kegg-categories}
kegg_gs_metadata <- read.csv("../Data/gene-set/kegg_geneset_metadata.csv", as.is = T)
kegg_categories <- factor(kegg_gs_metadata$category, levels = KEGG_TOP_CATEGORIES)
names(kegg_categories) <- kegg_gs_metadata$pathway_id
```

```{r reactome-categories}
react_gs_metadata <- read.csv("../Data/gene-set/reactome_geneset_metadata.csv", as.is = T)
react_categories <- factor(react_gs_metadata$category, levels = REACTOME_TOP_LEVEL_PATHWAYS)
names(react_categories) <- react_gs_metadata$pathway_id
```

```{r enriched-categories-kegg, fig.width=14}
data <- all_enrich_collapsed %>% dplyr::filter(db == "kegg", fdr < threshold)
categories <- kegg_categories[data$pathway_id]
kegg_count_categories <- plot.enriched.categories(data, categories, common_pathway_id)
kegg_count_categories
```

### Number of Enriched Reactome Categories

```{r enriched-categories-reactome, fig.width=14}
data <- all_enrich_collapsed %>% dplyr::filter(db == "reactome", fdr < threshold)
categories <- react_categories[data$pathway_id]
react_count_categories <- plot.enriched.categories(data, categories, common_pathway_id)
react_count_categories
```

# Commons Pathways

### Pathways Table

```{r table-common-pathways}
all_categ_enriched <- forcats::fct_c(kegg_categories, react_categories)
names(all_categ_enriched) <- c(names(kegg_categories), names(react_categories))
# Create a data table for display
common_pathway_2_go <- all_enrich_collapsed %>% 
  dplyr::filter(pathway_id %in% common_pathway_id) %>% 
  dplyr::distinct(pathway_id, description, db) %>%
  dplyr::left_join(id2go, by = c("pathway_id" = "source_id")) %>%
  dplyr::arrange(go_id) %>%
  dplyr::mutate(category = all_categ_enriched[pathway_id]) %>%
  dplyr::select(pathway_id, description, db, category, go_id, go_desc)

DT::datatable(
  common_pathway_2_go,
  rownames = F,
  caption = "Commons pathways with their mapped GO Terms"
)
```

```{r save-session, show=FALSE}
save.image("1_pdcl_enrichment.RData")
```
