---
title: "TCGA Personnalized"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)

# Load required libraries (others are load directly in sourced files)
library(Biobase)
library(dplyr)
library(purrr)
library(plotly)
library(DT)
library(mclust)
library(factoextra)
library(penda)

# Source files with functions used in next steps
source("R/gprofiler-analysis.R")
source("R/gsea-analysis.R")
source("R/process-result.R")
source("R/quickgo.R")
source("R/kegg.R")
source("R/reactome.R")

# Constant
rerun_rds_cached <- FALSE
result_dir <- "~/PhD/PDCL/Result_TCGA_personnalized/"
```

```{r prepare-count-data, eval=FALSE}
# The matrix was prepared beforehand, we just need to the txt file
count_data <- read.table("Data/TCGA-GBM-v32.0/count_matrix.txt", header = T, sep = "\t", as.is = T, row.names = 1, check.names = F)
count_data <- count_data[isUnique(count_data$gene_name),] # Keep unique genes only
row.names(count_data) <- count_data$gene_name
count_data <- count_data[,-1] # remove the gene names column
count_metadata <- read.table("Data/TCGA-GBM-v32.0/gdc_sample_sheet.2022-07-18.tsv", header = T, as.is = T, sep = "\t", check.names = F)
names(count_metadata) <- c("file_id",	"file_name", "data_category", "data_type", "project_id", "case_id", "sample_id", "sample_type")
count_metadata$condition <- "tumor"
index_normal <- grep("normal", count_metadata$sample_type, ignore.case = T, perl = T)
count_metadata$condition[index_normal] <- "solid_tissue_normal"
count_metadata$condition <- as.factor(count_metadata$condition)
```

```{r make-data}
dds <- DESeqDataSetFromMatrix(
  countData = count_data, 
  colData = count_metadata, 
  design = ~condition
)
dds <- DESeq(dds)
pld <- normTransform(dds)
normalized_count <- assay(pld, blind = F)
control_index <- which(count_metadata$condition == "control")
control_names <- colnames(normalized_count)[control_index]
all_data <- as.matrix(normalized_count)
```

<!-- 
  Here we prepare several PenDA datasets to perform crossed-validation.
  We don't have enough controls so we use this approach to ensure the quality
  of the Differential Expression analysis.
--> 
```{r make-penda-datasets}
nb_validation_run <- 10
validation_ctrl <- combn(control_names, 2, simplify = F)
validation_index <- sample(1:length(validation_ctrl), nb_validation_run)
validation_ctrl <- validation_ctrl[validation_index]

all_penda_datasets <- lapply(1:nb_validation_run, function(i){
  validation_ctrl_set <- validation_ctrl[[i]]
  names_ctrl_set <- control_names[!(control_names %in% validation_ctrl_set)]
  index_ctrl_set <- which(colnames(all_data) %in% names_ctrl_set)
  cases <- all_data[,-index_ctrl_set]
  ctrl <- all_data[,index_ctrl_set]
  penda_dataset <- penda::make_dataset(ctrl, cases, detectlowvalue = T, detectNA = T, threshold = 0.99)
})
```

```{r compute-L-H-lists}
all_L_H_list <- lapply(all_penda_datasets, function(penda_dataset){
  penda::compute_lower_and_higher_lists(penda_dataset$data_ctrl, threshold = 0.99, s_max = 30)
})
```

```{r run-penda}
threshold_test <- 0.6
iterations <- 20
quant_test <- 0.0
factor_test <- 1.2
all_penda_res <- lapply(1:length(all_penda_datasets), function(i){
  penda_dataset <- all_penda_datasets[[i]]
  L_H_list <- all_L_H_list[[i]]
  penda::penda_test(
    samples = penda_dataset$data_case, 
    controls = penda_dataset$data_ctrl, 
    threshold = threshold_test, 
    iterations = iterations,
    L_H_list = L_H_list,
    quant_test = quant_test,
    factor_test = factor_test
  )
})
```

<!-- Define GMT path -->

```{r define-gmt-path}
gmt_path <- list(
  reactome = "Data/gene-set/reactome_no_less_15_no_more_500.gmt",
  kegg = "Data/gene-set/kegg_no_less_15_no_more_500.gmt"
)
```

<!-- Load G:Profiler GMTs -->

<!-- Upload GMT File before running G:Profiler Enrichment Analysis if needed. Otherwise juste define the token of the GMT stored on G:Profiler website. -->

```{r create-gprofiler-gmt-list}
# Use RDS caching so we don't have to reupload the gmt files each times
gprofiler_gmt_list <- xfun::cache_rds({
  lapply(gmt_path, function(x){ upload_GMT_file(x) })
}, file = "gprofiler_gmt.rds", dir = result_dir, rerun = rerun_rds_cached)
```


```{r create-all-res-list}
all_res <- list()
```

```{r run-gprofiler}
all_pdcl_genes <- row.names(penda_res)
all_gene_list <- lapply(penda_res, function(x){
  all_pdcl_genes[x != 0]
})
# Use RDS caching as G:Profiler run are long for 20 samples
all_res[["G:Profiler"]] <- xfun::cache_rds({
  run.gost(all_gene_list, gprofiler_gmt_list, all_pdcl_genes)
}, file = "gprofiler_result", dir = result_dir, rerun = rerun_rds_cached)
```

```{r convert-to-enrichment} 
all_enrichment <- list()
all_enrichment[["G:Profiler"]] <- map_depth(all_res[["G:Profiler"]], .depth = 1, .f = convert.gost.to.enrichment)
```

<!-- Retrieve Categories for enriched pathways -->

```{r gs-metadata}
kegg_gs_metadata <- read.csv("Data/gene-set/kegg_geneset_metadata.csv", as.is = T)
react_gs_metadata <- read.csv("Data/gene-set/reactome_geneset_metadata.csv", as.is = T)
all_gs_metadata <- rbind(kegg_gs_metadata, react_gs_metadata)
```

```{r collapse-by-method}
# Collapse the results by method.
# One line contain the result of one enrichment method for one pathway of one PDCL
gs_category <- all_gs_metadata %>% dplyr::select(pathway_id, category)
all_enrich_collapsed <- collapse.enrichment(all_enrichment) %>% dplyr::inner_join(gs_category)
threshold <- 0.05
```

```{r filter-significant}
# We need to turn the category column as a factor so when we count the number
# of pathways per category, categories with 0 pathways are not dropped
kegg_significant <- all_enrich_collapsed %>%
  dplyr::filter(fdr < threshold, db == "kegg") 

reactome_significant <- all_enrich_collapsed %>%
  dplyr::filter(fdr < threshold, db == "reactome")
```

# Category Enriched {data-navmenu="Kegg"}

## {.tabset}

### Heatmap Categories

```{r nb-pathway-enriched-per-kegg-categories, fig.width=16, fig.height=10}
count <- kegg_significant %>%
  dplyr::mutate(category = factor(category, levels = KEGG_TOP_CATEGORIES)) %>% 
  dplyr::count(sample, method, category, .drop = F)

enriched_categories_kegg <- ggplot(
  data = count,
  aes(x = sample, y = category, fill = n)
) + 
  geom_raster() +
  labs(
    title = "Number of pathways enriched in each categories for each PDCL",
    x = "PDCL",
    y = "Categories",
    fill = "Count"
  ) +
  scale_fill_viridis_c(direction = 1) +
  theme(axis.text.x = element_text(angle = 90) ) +
  # By default it is sorted in desc order, we reverse it so it is in asc order
  scale_y_discrete(limits=rev)
enriched_categories_kegg
```

### Barplot Categories

```{r common-pathways-kegg, fig.width=14, fig.height=8}
rank_pdcl <- kegg_significant %>%
  dplyr::mutate(sample = factor(sample, levels = pdcl_names)) %>%
  dplyr::count(sample, .drop = FALSE) %>%
  dplyr::arrange(desc(n)) %>%
  dplyr::pull(sample)

data <- kegg_significant %>%
  dplyr::mutate(sample = factor(sample, levels = rank_pdcl))

barplot_kegg_categories <- ggplot(data, aes(x = sample, fill = category)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(angle = 90) )
barplot_kegg_categories
```

# Enrichment Per PDCL {data-navmenu="Kegg"}

## {.tabset}

### Heatmap FDR Metabolism and Cellular Processes

```{r heatmap-fdr-kegg-metabolism-cellular-processes, fig.width=18, fig.height=12}
data <- prepare.fdr.data(kegg_significant$pathway_id, "\\bmetabolism|cellular\\s+processes\\b")
heatmap_fdr_kegg_metabolism_ecm <- heatmap.fdr(data)
heatmap_fdr_kegg_metabolism_ecm
```

### Heatmap FDR Cell Cycle

```{r heatmap-fdr-kegg-organismal-systems, fig.width=18, fig.height=12}
data <- prepare.fdr.data(kegg_significant$pathway_id, "organismal\\s+systems")
heatmap_fdr_kegg_cell_cycle_signal <- heatmap.fdr(data)
heatmap_fdr_kegg_cell_cycle_signal
```

# Frequently Deregulated {data-navmenu="Kegg"}

### Table of the frequently deregulated genes

```{r freq-dereg-kegg}
freq_dereg_kegg <- kegg_significant %>%
  dplyr::count(pathway_id, description, category) %>%
  dplyr::arrange(desc(n)) %>%
  dplyr::rename(count=n)
datatable(freq_dereg_kegg)
```

# Clustering {data-navmenu="Kegg" data-orientation=columns}

```{r count-tot-for-cluster}
gmt <- lapply(gmt_path, read.gmt)
list_tot_count <- purrr::imap(gmt, function(x,y){
  all_gs_metadata %>% 
    dplyr::filter(pathway_id %in% names(x)) %>%
    dplyr::count(category) %>%
    dplyr::mutate(category = factor(category))
})
```


```{r cluster-kegg}
table_kegg_category <- kegg_significant %>% 
  dplyr::select(sample, category) %>%
  dplyr::mutate(category = factor(category, levels = list_tot_count$kegg$category)) %>%
  dplyr::count(sample, category, .drop = F) %>%
  tidyr::pivot_wider(names_from = category, values_from = n) %>%
  tibble::as_tibble()
clustering_vector_kegg <- mapply("/", table_kegg_category[-1], list_tot_count$kegg$n)
row.names(clustering_vector_kegg) <- table_kegg_category$sample
clustering_vector_kegg <- clustering_vector_kegg[,colSums(clustering_vector_kegg) > 0]
# Other way to do it
# clustering_vector_kegg <- sweep(clustering_vector_kegg, 2, list_tot_count$kegg$n, "/")
kmeans_kegg <- kmeans(clustering_vector_kegg, centers = 5, nstart = 50)
mclust_kegg <- mclust::Mclust(clustering_vector_kegg)
pca_kegg <- prcomp(clustering_vector_kegg, scale = T)
```

## {.tabset}

### Cluster Plot K-Means KEGG

```{r cluster-plot-kmeans-kegg, fig.width=12, fig.height=7}
cluster_plot_kmeans_kegg <- fviz_cluster(kmeans_kegg, clustering_vector_kegg, repel = TRUE) + 
  labs(title = "Cluster Plot K-Means KEGG (5 centers)") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha=0.6) + 
  geom_vline(xintercept = 0, linetype = "dashed", alpha=0.6)
cluster_plot_kmeans_kegg
```

### Cluster Plot MClust KEGG

```{r cluster-plot-mclust-kegg, fig.width=12, fig.height=7}
cluster_plot_mclust_kegg <- fviz_cluster(mclust_kegg, clustering_vector_kegg, repel = TRUE) +
  labs(title = "Cluster Plot MClust KEGG") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha=0.6) + 
  geom_vline(xintercept = 0, linetype = "dashed", alpha=0.6)
cluster_plot_mclust_kegg
```

##

### Var Correlation

```{r plot-corr-circle-kegg, fig.width=8, fig.height=8}
corr_circle_plot_kegg <- fviz_pca_var(pca_kegg, axes = c(1,2), repel = T)
corr_circle_plot_kegg
```


# Category Enriched {data-navmenu="Reactome"}

## {.tabset}

### Heatmap Categories

```{r nb-pathway-enriched-per-reactome-categories, fig.width=16, fig.height=10}
count <- reactome_significant %>%
  dplyr::mutate(category = factor(category, levels = REACTOME_TOP_LEVEL_PATHWAYS)) %>%
  dplyr::count(sample, method, category, .drop = F)

enriched_categories_reactome <- ggplot(
  data = count,
  aes(x = sample, y = category, fill = n)
) + 
  geom_raster() +
  labs(
    title = "Number of pathways enriched in each categories for each PDCL",
    x = "PDCL",
    y = "Categories",
    fill = "Count"
  ) +
  scale_fill_viridis_c(direction = 1) +
  theme(axis.text.x = element_text(angle = 90) ) +
  # By default it is sorted in desc order, we reverse it so it is in asc order
  scale_y_discrete(limits=rev)
enriched_categories_reactome
```

### Barplot Categories

```{r common-pathways-reactome, fig.width=14, fig.height=8}
rank_pdcl <- reactome_significant %>%
  dplyr::mutate(sample = factor(sample, levels = pdcl_names)) %>%
  dplyr::count(sample, .drop = FALSE) %>%
  dplyr::arrange(desc(n)) %>%
  dplyr::pull(sample)

data <- reactome_significant %>%
  dplyr::mutate(sample = factor(sample, levels = rank_pdcl))

barplot_reactome_categories <- ggplot(data, aes(x = sample, fill = category)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(angle = 90) )
barplot_reactome_categories
```

# Enrichment Per PDCL {data-navmenu="Reactome"}

## {.tabset}

### Heatmap FDR Metabolism and ECM

```{r heatmap-fdr-reactome-metabolism-ecm, fig.width=18, fig.height=12}
data <- prepare.fdr.data(reactome_significant$pathway_id, "\\bmetabolism|matrix\\b")
heatmap_fdr_reactome_metabolism_ecm <- heatmap.fdr(data)
heatmap_fdr_reactome_metabolism_ecm
```

### Heatmap FDR Cell Cycle

```{r heatmap-fdr-reactome-cell-cycle-and-signalling, fig.width=18, fig.height=12}
data <- prepare.fdr.data(reactome_significant$pathway_id, "\\b(cell.*|dna|signal|processing)\\b")
heatmap_fdr_reactome_cell_cycle_signal <- heatmap.fdr(data)
heatmap_fdr_reactome_cell_cycle_signal
```

# Frequently Deregulated {data-navmenu="Reactome"}

### Table of the frequently deregulated genes

```{r freq-dereg-reactome}
freq_dereg_reactome <- reactome_significant %>%
  dplyr::count(pathway_id, description, category) %>%
  dplyr::arrange(desc(n)) %>%
  dplyr::rename(count=n)
datatable(freq_dereg_reactome)
```

# Clustering {data-navmenu="Reactome" data-orientation=columns}

```{r cluster-reactome}
table_reactome_category <- reactome_significant %>% 
  dplyr::select(sample, category) %>%
  dplyr::mutate(category = factor(category, levels = list_tot_count$reactome$category)) %>%
  dplyr::count(sample, category, .drop = F) %>%
  tidyr::pivot_wider(names_from = category, values_from = n) %>%
  tibble::as_tibble()
clustering_vector_reactome <- mapply("/", table_reactome_category[-1], list_tot_count$reactome$n)
row.names(clustering_vector_reactome) <- table_reactome_category$sample
clustering_vector_reactome <- clustering_vector_reactome[,colSums(clustering_vector_reactome) > 0]
# Other way to do it
# clustering_vector_kegg <- sweep(clustering_vector_kegg, 2, list_tot_count$kegg$n, "/")
kmeans_reactome <- kmeans(clustering_vector_reactome, centers = 3, nstart = 50)
mclust_reactome<- mclust::Mclust(clustering_vector_reactome)
pca_reactome <- prcomp(clustering_vector_reactome, scale = T)
```

## {.tabset}

### Cluster Plot K-Means reactome

```{r cluster-plot-kmeans-reactome, fig.width=12, fig.height=7}
cluster_plot_kmeans_reactome <- fviz_cluster(kmeans_reactome, clustering_vector_reactome, repel = TRUE) + 
  labs(title = "Cluster Plot K-Means reactome (5 centers)") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha=0.6) + 
  geom_vline(xintercept = 0, linetype = "dashed", alpha=0.6)
cluster_plot_kmeans_reactome
```

### Cluster Plot MClust reactome

```{r cluster-plot-mclust-reactome, fig.width=12, fig.height=7}
cluster_plot_mclust_reactome <- fviz_cluster(mclust_reactome, clustering_vector_reactome, repel = TRUE) +
  labs(title = "Cluster Plot MClust reactome")+
  geom_hline(yintercept = 0, linetype = "dashed", alpha=0.6) + 
  geom_vline(xintercept = 0, linetype = "dashed", alpha=0.6)
cluster_plot_mclust_reactome
```

##

### Var Correlation

```{r plot-corr-circle-reactome, fig.width=8, fig.height=8}
corr_circle_plot_reactome <- fviz_pca_var(pca_reactome, axes = c(1,2), repel = T)
corr_circle_plot_reactome
```

# Count Enriched Pathways

### Count Enriched pathways per Databases

```{r count-enriched-db, fig.width=14, fig.height=8}
data <- all_enrich_collapsed %>% dplyr::filter(fdr<threshold)
count_enriched_method <- ggplot(
  data = data,
  aes(x = sample, fill = db)
) +
  geom_bar(position = "dodge2") +
  labs(
    title = "Number of enriched pathways per databases",
    x = "PDCL sample",
    y = "Counts",
    fill = ""
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  theme(axis.text.x = element_text(angle = 90) )
count_enriched_method
```

```{r plot-contrib-pca-kegg, fig.width=8, fig.height=8, eval=FALSE}
plot_contrig_pca_kegg <- fviz_contrib(pca_kegg, choice = "var")
plot_contrig_pca_kegg
```


```{r save-session}
save.image("~/PhD/PDCL/pdcl_process_personalized.RData")
```
