---
title: "Cross Validation Penda on PDCL"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
# Load required libraries (others are load directly in sourced files)
library(Biobase)
library(dplyr)
library(purrr)
library(plotly)
library(DT)
library(mclust)
library(factoextra)
library(penda)
library(DESeq2)

# Source files with functions used in next steps
source("../Code/R/pdcl-dataset.R")

validation_datasets_param <- list(
  nb_validation_run = 10,
  nb_validation_ctrl = 2
)
penda_datasets_param <- list(
  detectlowvalue = T, 
  detectNA = T, 
  threshold = 0.99
)
penda_L_H_params <- list(
  threshold = 0.99, 
  s_max = 130
)
penda_test_param <- list(
  threshold_test = 0.45,
  iterations = 10,
  quant_test = 0.0,
  factor_test = 1.6
)
rerun_cache <- T 
result_dir = "Result_pdcl_cross_validation/"
```

# Dataset and Data Filtering

## Prepare Count-Data
```{r prepare-count-data}
control_count_data <- read.csv(PDCL_DATASET_PARAM$control_count_path)
pdcl_count_data <- read.csv(PDCL_DATASET_PARAM$pdcl_count_path, check.names = F)
# Merge control and PDCL counts into one dataframe
count_data <- merge(control_count_data, pdcl_count_data, by.x = "symbol", by.y = "gene_id")
# Associate each row with a gene symbol
row.names(count_data) <- count_data$symbol
# Remove symbol and id columns
count_data <- count_data %>% dplyr::select(!c(symbol, id))
condition_vect <- c(
  rep("control", length(PDCL_DATASET_PARAM$control_sample_names)), 
  rep("pdcl", length(PDCL_DATASET_PARAM$pdcl_sample_names))
)
sample <- gsub("^(.*)_(Br\\d.*)", "\\1", colnames(count_data), ignore.case = T, perl = T, fixed = F)
# sample[(length(control_names)+1):length(sample)] <- "gbm"
replicate <- gsub("^(.*)_(Br\\d.*)", "\\2", colnames(count_data), ignore.case = T, perl = T, fixed = F)
count_metadata <- data.frame(
  sample = as.factor(sample),
  replicate = as.factor(replicate),
  condition = as.factor(condition_vect),
  row.names = colnames(count_data)
)
```

## Pseudo-log Normalization

```{r make-data}
dds <- DESeqDataSetFromMatrix(
  countData = count_data, 
  colData = count_metadata, 
  design = ~condition
)
dds <- DESeq(dds)
pld <- normTransform(dds)
normalized_count <- assay(pld, blind = F)
control_index <- which(count_metadata$condition == "control")
control_names <- colnames(normalized_count)[control_index]
all_data <- as.matrix(normalized_count)
```

## Sample Validation

PCA visualization to compare controls and tumour samples to ensure that they can be classified in 2 groups
and that controls are not too different from each others.

```{r pca-visualization, echo=FALSE}
t_data <- t(all_data)
t_data <- t_data[, colSds(t_data) != 0]
pca_obj <- prcomp(t_data, scale. = T)
pca_plot <- fviz_pca_ind(pca_obj, col.ind = count_metadata$condition, repel = T)
pca_plot
```

Density of the normalized counts using the Pseudo-Log transformation from DESeq2 to compare the distribution
of counts.

```{r density-visualization, echo=FALSE}
all_density <- purrr::map_dfr(1:ncol(all_data), function(i){
  density_obj <- density(all_data[,i]) 
  data.frame(count = density_obj$x, density = density_obj$y, sample_index = i)
})
all_density$sample_type <- count_metadata$sample_type[all_density$sample_index]
all_density$condition <- count_metadata$condition[all_density$sample_index]
density_plot <- ggplot(all_density, aes(count, density, group = sample_index, colour = condition)) +
  geom_line(show.legend = F) +
  facet_wrap(~condition, ncol = 1) +
  labs(title = "Density plot of the log(count+1)")
density_plot
```


## Data filtering

Here we prepare several PenDA datasets to perform crossed-validation.
We don't have enough controls so we use this approach to ensure the quality of the Differential Expression analysis.

```{r make-penda-datasets}
validation_ctrl <- combn(control_names, validation_datasets_param$nb_validation_ctrl, simplify = F)
validation_index <- sample(1:length(validation_ctrl), validation_datasets_param$nb_validation_run)
validation_ctrl <- validation_ctrl[validation_index]

all_penda_datasets <- xfun::cache_rds({
  with(penda_datasets_param, {
    lapply(1:validation_datasets_param$nb_validation_run, function(i){
      validation_ctrl_set <- validation_ctrl[[i]]
      names_ctrl_set <- control_names[!(control_names %in% validation_ctrl_set)]
      index_ctrl_set <- which(colnames(all_data) %in% names_ctrl_set)
      cases <- all_data[,-index_ctrl_set]
      ctrl <- all_data[,index_ctrl_set]
      penda::make_dataset(ctrl, cases, detectlowvalue = detectlowvalue, detectNA = detectNA, threshold = threshold)
    })
  })
}, rerun = rerun_cache, file = "all_penda_datasets.rds", dir = result_dir)
```

## Relative gene ordering

```{r compute-L-H-lists}
all_L_H_list <- with(penda_L_H_params, {
  lapply(all_penda_datasets, function(penda_dataset){
    penda::compute_lower_and_higher_lists(penda_dataset$data_ctrl, threshold = threshold, s_max = s_max)
  })
})
```


# Differential Expression Analysis

# Differential Expression testing

We hide the results of this chunck because it will overload the doc with unnecessary comments.

```{r run-penda, results='hide'}
all_penda_res <- xfun::cache_rds({
  with(penda_test_param, {
    lapply(1:length(all_penda_datasets), function(i){
      penda_dataset <- all_penda_datasets[[i]]
      L_H_list <- all_L_H_list[[i]]
      penda::penda_test(
        samples = penda_dataset$data_case, 
        controls = penda_dataset$data_ctrl, 
        threshold = threshold_test, 
        iterations = iterations,
        L_H_list = L_H_list,
        quant_test = quant_test,
        factor_test = factor_test
      )
    })
  })
}, rerun = rerun_cache, file = "all_penda_res.rds", dir = result_dir)
```

## Common Deregulations

```{r common-dereg-percentage, echo = FALSE, fig.width=12, fig.height=8}
results_combination <- combn(1:length(all_penda_res), 2, simplify = F)
common_dereg_percentage <- purrr::imap_dfr(results_combination, function(combination, i){
  index_res1 <- combination[1]
  index_res2 <- combination[2]
  res1 <- all_penda_res[[index_res1]]
  res2 <- all_penda_res[[index_res2]]
  if( any(dim(res1$down_genes) != dim(res2$down_genes)) | any(dim(res1$up_genes) != dim(res2$up_genes)) ){
    warning("Down or Up genes matrix dimensions don't match between ", index_res1, " and ", index_res2, ". Returning 0% commons deregulation.")
    return( data.frame(sample = colnames(res1$down), down = 0, up = 0, combination_numb = i, combination_pair = paste0(combination, collapse = "-"), row.names = NULL) )
  }
  res1 <- lapply(all_penda_res[[index_res1]], function(x){
    x[, !(colnames(x) %in% control_names)]
    x <- x[order(row.names(x)),]
    x
  })
  res2 <- lapply(all_penda_res[[index_res2]], function(x){
    x[, !(colnames(x) %in% control_names)]
    x <- x[order(row.names(x)),]
    x
  })
  down_commons <- colSums(res1$down & res2$down)/colSums(res1$down | res2$down)
  up_commons <- colSums(res1$up & res2$up)/colSums(res1$up | res2$up)
  data.frame(sample = colnames(res1$down), down=down_commons, up = up_commons, combination_numb = i, combination_pair = paste0(combination, collapse = "-"), row.names = NULL)
})

data <- common_dereg_percentage %>%
  dplyr::filter(!(sample %in% control_names)) %>%
  tidyr::pivot_longer(tidyr::matches("up|down|total", ignore.case = T), names_to = "dereg", values_to = "percent") %>%
  dplyr::mutate(run = as.factor(combination_numb))

plot_common_dereg_percent <- ggplot(data, aes(x = run, y = percent, colour = dereg)) +
  geom_point() +
  geom_hline(yintercept = 0.97, colour = "red", alpha = 0.6) 

plot_common_dereg_percent
```

```{r dereg-perecent, fig.width=12, fig.height=8}
dereg_percentage <- purrr::imap_dfr(all_penda_res, function(penda_res, run){
  up <- colSums(penda_res$up_genes)/nrow(penda_res$up_genes)
  down <- colSums(penda_res$down_genes)/nrow(penda_res$down_genes)
  total <- up + down
  data.frame(sample = names(up), up = up, down = down, total = total, run = run, row.names = NULL)
})
```

## Deregulation Tumour Percent

```{r tumour-dereg-perecent, echo = FALSE}
data <- dereg_percentage %>%
  tidyr::pivot_longer(tidyr::matches("up|down|total", ignore.case = T), names_to = "dereg", values_to = "percent") %>%
  dplyr::filter(!(sample %in% control_names)) %>%
  dplyr::mutate(run = as.factor(run))

plot_dereg_percent_tumour <- ggplot(data, aes(x = run, y = percent, colour = dereg)) +
  geom_point() +
  geom_hline(yintercept = c(0.05, 0.1, 0.3, 0.5), colour = "grey", linetype = "dashed") +
  ylim(c(0,1))
plot_dereg_percent_tumour
```

## Deregulation Controls Percentage

```{r controls-dereg-perecent, echo = FALSE}
data <- dereg_percentage %>%
  tidyr::pivot_longer(tidyr::matches("up|down|total", ignore.case = T), names_to = "dereg", values_to = "percent") %>%
  dplyr::filter(sample %in% control_names) %>%
  dplyr::mutate(run = as.factor(run))

plot_dereg_percent_control <- ggplot(data, aes(x = run, y = percent, colour = dereg)) +
  geom_point() +
  geom_hline(yintercept = c(0.05, 0.1), colour = c("red", "blue"), linetype = "dotted") +
  ylim(c(0,1))
plot_dereg_percent_control
```

# Run information

## Parameter Values
```{r print-param}
validation_datasets_param
penda_datasets_param
penda_L_H_params
penda_test_param
rerun_cache
result_dir
```

## Session Information

```{r session-info}
sessionInfo()
```

```{r save-image, include=FALSE}
save.image("pdcl_cross_validation.RData")
```
